---
title: "Tea Blending Optimisation"
subtitle: "STA5074Z Prescriptive Analytics"
author: "Tinotenda Muzambi - MZMTIN002"
date: "`r Sys.Date()`"
bibliography: references.bib
csl: apa-numeric-superscript-brackets.csl
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 5, fig.height = 3, fig.align = "center")

# Load required libraries
library(dplyr)
library(Rglpk)
library(ompr)
library(ompr.roi)
library(ROI)
library(ROI.plugin.glpk)
library(lpSolveAPI)
library(ggplot2)
library(GA)
library(knitr)
library(flextable) 
library(tidyr)

# Helper function to print tables 
print.table <- function(data, caption = "", font.size = 9) {
  # Round all numeric columns to 2 decimal places
  data.rounded <- data
  for (col in names(data.rounded)) {
    if (is.numeric(data.rounded[[col]])) {
      data.rounded[[col]] <- round(data.rounded[[col]], 2)
    }
  }
  
  # Determine if table is wide
  is.wide <- ncol(data.rounded) > 8
  
  ft <- flextable(data.rounded) %>%
    set_caption(caption) %>%
    theme_zebra(odd_header = "#CFCFCF", odd_body = "#EFEFEF")
  
  if (is.wide) {
    # For wide tables, use smaller font
    ft <- ft %>%
      fontsize(size = font.size) %>%
    autofit()
  } else {
    ft <- ft %>%
      fontsize(size = font.size) %>%
      autofit()
  }
  
  if (knitr::is_latex_output()) {
    ft <- ft %>%
      set_table_properties(layout = "autofit")
  } else {
    # For HTML output
    if (is.wide) {
      # For wide tables, ensure they fit and can scroll
      ft <- ft %>%
        set_table_properties(layout = "autofit", width = 1)
    } else {
      ft <- ft %>%
      set_table_properties(layout = "autofit")
    }
  }
  
  return(ft)
}
```

<style>
/* Ensure wide tables scroll horizontally */
.ft-container {
  overflow-x: auto;
  max-width: 100%;
}
table {
  max-width: 100%;
}
</style>

\newpage

# 1 Introduction
Blending problems are a classic challenge in operations research, arising in diverse fields such as food processing, metallurgy, and energy. In the tea industry, blending is crucial to achieve consistent product quality, balance raw material costs, and adapt to fluctuating supplies. Every cup of tea on the supermarket shelf is the result of a carefully orchestrated process. Different grades and types of tea leaves from various origins are combined to meet precise targets for taste, colour, aroma and other quality parameters, all while minimising production costs and ensuring supply constraints are not violated.

In the context of tea manufacturing, the blending optimisation problem involves selecting the optimal composition of raw teas to satisfy desired blend qualities for multiple final blends. This not only impacts the sensory profile and brand identity of each product, but also has substantial operational and economic implications. By applying advanced computational techniques, such as mathematical programming and metaheuristics, producers can automate and enhance the blending decision-making process, yielding better margins, consistent quality and greater responsiveness to market demands.

Below, we formally define the tea blending optimisation problem and systematically explore several approaches for solving it. The problem requires determining the least-cost combination of raw materials to produce final blends that meet specific quality characteristics, as formalised by Fomeni [@fomeni].

The primary objectives are:

1. To implement and solve a cost-minimisation model using Linear Programming (LP).
2. To solve the same problem using Simulated Annealing (SA) metaheuristic and compare performance.
3. To formulate and solve a multi-objective model using Chebyshev Goal Programming.
4. To explore Genetic Algorithms (GAs) as an alternative solution approach.

# 2 Problem Formulation and Data Preparation

## 2.1 Notations and Parameters

- **Indices**:
  - $i$: Index for raw materials
  - $j$: Index for final blends
  - $k$: Index for quality characteristics

- **Parameters**:
  - $P_i$: Unit cost of raw material $i$ (£/Kg)
  - $D_j$: Weekly demand for final blend $j$ (Kgs)
  - $a_i$: Weekly availability of raw material $i$ (Kgs)
  - $g_{ik}$: Grade of characteristic $k$ in one unit of raw material $i$
  - $s_{jk}$: Required target score of characteristic $k$ in one unit of final blend $j$

- **Decision Variables**:
  - $x_{ij}$: Amount (in Kgs) of raw material $i$ to be used in final blend $j$

## 2.2 Data Source

This project uses the real-world data from Fomeni [@fomeni], extracted from tables 2 and 3 of the paper. These data represent actual tea blending scenarios from a UK-based tea company. The data are loaded and prepared in section 3 below.

# 3 Data Implementation (Fomeni 2018)

```{r init-vars, include=FALSE}
# Initialise basic counts for inline R code
n.materials.fomeni <- 60
n.blends.fomeni <- 10
n.characteristics.fomeni <- 6
```

This section implements the tea blending optimisation problem using the actual data provided in Fomeni [@fomeni]. 

## 3.1 Data Loading

The dataset comprises **`r n.materials.fomeni` raw materials** and **`r n.blends.fomeni` final blends**, each characterised by **`r n.characteristics.fomeni` quality attributes**, C1 through C6.

```{r fomeni-data, echo=FALSE}
# Real data from Fomeni (2018) - Table 2: Raw Materials
# Note: Prices are in £0.01/kg so we convert to £/kg by dividing by 100
# Quality characteristics: C1, C2, C3, C4, C5, C6

raw.materials.fomeni <- data.frame(
  Material.ID = paste0("R", 1:60),
  C1 = c(5.0, 5.0, 5.0, 5.0, 5.0, 7.0, 7.0, 8.0, 8.0, 8.0, 2.0, 2.0, 2.0, 3.0, 4.0, 4.0, 7.0, 4.0, 4.0, 5.0, 5.0, 5.0, 5.0, 6.0, 6.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 9.0, 2.0, 5.0, 5.0, 5.0, 5.0, 2.0, 4.0, 5.0, 5.0, 2.0, 5.0, 5.0, 5.0, 6.0, 6.0, 7.0, 8.0, 8.0, 5.0, 5.0, 5.0, 7.0, 8.0, 8.0, 2.0, 4.0, 6.0, 6.0),
  C2 = c(4.0, 3.0, 4.0, 3.0, 4.0, 4.0, 4.0, 4.0, 4.0, 4.0, 2.0, 2.0, 0.5, 0.5, 0.5, 2.0, 0.5, 8.0, 8.0, 6.0, 6.0, 6.0, 6.0, 6.0, 6.0, 5.0, 5.0, 5.0, 5.0, 5.0, 5.0, 5.0, 0.5, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 3.0, 3.0, 3.0, 3.0, 3.0, 2.0, 2.0, 0.5, 1.0, 1.0, 1.0, 2.0, 8.0, 5.0, 5.0),
  C3 = c(9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 3.0, 3.0, 1.0, 1.0, 1.0, 3.0, 1.0, 5.0, 5.0, 4.0, 4.0, 4.0, 4.0, 4.0, 4.0, 4.0, 6.0, 6.0, 7.0, 6.0, 6.0, 6.0, 1.0, 2.0, 2.0, 2.0, 3.0, 3.0, 3.0, 2.0, 3.0, 5.0, 4.0, 4.0, 4.0, 4.0, 4.0, 3.0, 3.0, 3.0, 2.0, 2.0, 1.0, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0),
  C4 = c(6.0, 4.0, 6.0, 4.0, 6.0, 6.0, 6.0, 6.0, 6.0, 6.0, 2.0, 2.0, 5.0, 5.0, 5.0, 2.0, 5.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 5.0, 3.0, 3.0, 3.0, 3.0, 2.0, 2.0, 3.0, 3.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 3.0, 3.0, 3.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0),
  C5 = c(5.0, 5.0, 5.0, 5.0, 5.0, 7.0, 7.0, 8.0, 8.0, 8.0, 2.0, 2.0, 2.0, 3.0, 4.0, 4.0, 7.0, 4.0, 4.0, 5.0, 5.0, 5.0, 5.0, 6.0, 6.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 9.0, 2.0, 5.0, 5.0, 5.0, 5.0, 2.0, 4.0, 5.0, 5.0, 2.0, 5.0, 5.0, 5.0, 6.0, 6.0, 7.0, 8.0, 8.0, 5.0, 5.0, 5.0, 7.0, 8.0, 8.0, 2.0, 4.0, 6.0, 6.0),
  C6 = c(5.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 5.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 9.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 5.0, 1.0, 5.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 5.0, 1.0, 5.0, 1.0, 5.0, 5.0, 1.0, 5.0, 1.0, 1.0, 5.0, 5.0, 1.0, 1.0, 1.0, 1.0, 5.0),
  Price.001.per.kg = c(2.40, 1.67, 2.35, 1.70, 2.30, 2.65, 2.65, 2.94, 2.84, 2.84, 1.30, 1.30, 1.20, 1.70, 1.50, 2.50, 1.80, 1.30, 1.30, 1.80, 1.90, 1.80, 1.50, 1.90, 1.80, 2.60, 2.00, 2.00, 2.10, 2.00, 1.70, 2.40, 1.40, 1.60, 1.60, 1.60, 1.60, 1.30, 2.50, 1.60, 1.70, 2.10, 2.40, 2.50, 2.50, 2.70, 2.80, 3.30, 3.40, 3.60, 1.80, 1.70, 2.10, 2.90, 4.70, 4.20, 1.30, 1.60, 1.70, 4.00),
  Availability.kg = c(5000, 1000, 25000, 1000, 25000, 20000, 20000, 1000, 15000, 15000, 40000, 80000, 60000, 5000, 5000, 20000, 20000, 20000, 20000, 40000, 40000, 20000, 5000, 1000, 500, 1000, 40000, 40000, 25000, 30000, 5000, 5000, 10000, 10000, 10000, 10000, 10000, 20000, 15000, 2000, 5000, 1000, 5000, 5000, 5000, 2000, 2000, 1000, 1000, 1000, 1000, 5000, 5000, 1000, 1000, 1000, 5000, 40000, 5000, 1000),
  stringsAsFactors = FALSE
)

# Convert price from £0.01/kg to £/kg
raw.materials.fomeni$Cost <- raw.materials.fomeni$Price.001.per.kg / 100
raw.materials.fomeni$Availability <- raw.materials.fomeni$Availability.kg

# Extract quality characteristics matrix
char.names.fomeni <- c("C1", "C2", "C3", "C4", "C5", "C6")
quality.matrix.fomeni <- as.matrix(raw.materials.fomeni[, char.names.fomeni])

n.materials.fomeni <- nrow(raw.materials.fomeni)
n.characteristics.fomeni <- length(char.names.fomeni)
P.fomeni <- raw.materials.fomeni$Cost 

# Table 3: Blends data
blends.fomeni <- data.frame(
  Blend.ID = paste0("B", 1:10),
  C1 = c(2.30, 3.80, 6.60, 6.20, 4.70, 8.40, 4.00, 5.90, 7.00, 4.00),
  C2 = c(3.00, 4.70, 5.20, 5.20, 5.10, 4.70, 0.50, 0.50, 3.90, 2.00),
  C3 = c(3.00, 2.90, 6.10, 5.30, 4.50, 6.70, 1.00, 1.00, 9.00, 3.00),
  C4 = c(2.00, 3.00, 6.90, 6.40, 5.00, 7.30, 5.00, 3.90, 5.90, 2.00),
  C5 = c(2.32, 3.80, 6.60, 6.20, 4.70, 8.40, 4.00, 5.90, 7.00, 4.00),
  C6 = c(1.00, 1.00, 1.00, 1.00, 1.00, 1.00, 1.00, 1.00, 1.00, 9.00),
  Demand.kg = c(100000, 100000, 60000, 100000, 100000, 50000, 10000, 5000, 10000, 20000),
  stringsAsFactors = FALSE
)

blends.fomeni$Demand <- blends.fomeni$Demand.kg

n.blends.fomeni <- nrow(blends.fomeni)

# Extract target quality matrix
target.matrix.fomeni <- as.matrix(blends.fomeni[, char.names.fomeni])

# Display first few rows of raw materials
print.table(raw.materials.fomeni[1:10, c("Material.ID", "Cost", "Availability", char.names.fomeni)], 
            caption = "Raw Materials: Costs (£/kg), Availability (kg), and Quality Characteristics (First 10)")

# Display all blends
print.table(blends.fomeni[, c("Blend.ID", "Demand", char.names.fomeni)], 
            caption = "Blends: Demand (kg) and Target Quality Scores")
```

Raw material prices range from £`r sprintf("%.4f", min(P.fomeni))` to £`r sprintf("%.4f", max(P.fomeni))` per kilogram, with a mean price of £`r sprintf("%.4f", mean(P.fomeni))` per kilogram. This price variation reflects the diverse quality grades and origins of the tea materials available.

```{r fomeni-data-cont, echo=FALSE}
# Prepare data vectors for optimisation models

# Availability vector a[i]
a.fomeni <- raw.materials.fomeni$Availability

# Demand vector D[j]
D.fomeni <- blends.fomeni$Demand

# Quality matrices
g.fomeni <- quality.matrix.fomeni  # g[i,k]
s.fomeni <- target.matrix.fomeni   # s[j,k]

# Set tolerance parameter (epsilon)
epsilon.fomeni <- 0.5

# Feasibility check
total.availability.fomeni <- sum(a.fomeni)
total.demand.fomeni <- sum(D.fomeni)

# Standard variable names for use in optimisation models
n.materials <- n.materials.fomeni
n.blends <- n.blends.fomeni
n.characteristics <- n.characteristics.fomeni
char.names <- char.names.fomeni
P <- P.fomeni
a <- a.fomeni
D <- D.fomeni
g <- g.fomeni
s <- s.fomeni
epsilon <- epsilon.fomeni
raw.materials <- raw.materials.fomeni
blends <- blends.fomeni

# Create named versions for display
P.named <- P.fomeni
names(P.named) <- raw.materials.fomeni$Material.ID
a.named <- a.fomeni
names(a.named) <- raw.materials.fomeni$Material.ID
D.named <- D.fomeni
names(D.named) <- blends.fomeni$Blend.ID
g.named <- g.fomeni
rownames(g.named) <- raw.materials.fomeni$Material.ID
s.named <- s.fomeni
rownames(s.named) <- blends.fomeni$Blend.ID
```

A feasibility analysis reveals that the total raw material availability is **`r format(total.availability.fomeni, big.mark = " ")` kg**, while total demand across all blends is **`r format(total.demand.fomeni, big.mark = " ")` kg**. This represents a **`r sprintf("%.2f", (total.availability.fomeni - total.demand.fomeni) / total.demand.fomeni * 100)`% surplus** in availability, indicating that supply constraints are not binding and the problem has sufficient flexibility for optimisation. The surplus availability provides the solver with multiple feasible solution paths, which is advantageous for finding cost-optimal allocations.

# 4 Approach 1: Cost Minimisation using Linear Programming

This section implements the cost-minimisation model using Linear Programming with parametric relaxation of the characteristic-matching constraints. This involves implementing a cost-minimisation LP model that finds the optimal combination of raw materials to produce blends meeting demand and quality constraints. This implementation uses Rglpk which provides a straightforward matrix-based formulation that is well-suited for this problem type.

## 4.1 Model Formulation

**Objective Function**: Minimise total cost
$$\min \sum_{i} \sum_{j} P_i x_{ij}$$

**Subject to Constraints**:

1. Raw Material Availability: $\sum_{j} x_{ij} \le a_i \quad \forall i$
2. Blend Demand: $\sum_{i} x_{ij} \ge D_j \quad \forall j$
3. Characteristic Score Relaxation:
   - $\sum_{i} (g_{ik} - (s_{jk} + \epsilon_{jk})) x_{ij} \le 0 \quad \forall j, k$
   - $\sum_{i} (g_{ik} - (s_{jk} - \epsilon_{jk})) x_{ij} \ge 0 \quad \forall j, k$
4. Non-negativity: $x_{ij} \ge 0 \quad \forall i, j$

## 4.2 Implementation

The Linear Programming model was successfully solved using the Rglpk solver. The solver converged to an optimal solution, indicating that the problem is feasible and the global optimum has been identified.

```{r lp-implementation, echo=FALSE}
# Implementation using Rglpk

# Number of decision variables: n.materials * n.blends (x[i,j] for all i,j)
n.vars <- n.materials * n.blends

# Objective function: minimise cost
# Objective coefficients: P[i] for each x[i,j]
# Variable ordering: x[1,1], x[1,2], ..., x[1,n.blends], x[2,1], ..., x[n.materials,n.blends]
obj.coef <- rep(P, n.blends)  # Repeat P for each blend

# Build constraint matrix
# Constraints in order:
# 1. Availability: sum_j x[i,j] <= a[i] for each i (n.materials constraints)
# 2. Demand: sum_i x[i,j] >= D[j] for each j (n.blends constraints)  
# 3. Quality upper: sum_i (g[i,k] - (s[j,k] + eps)) * x[i,j] <= 0 (n.blends * n.characteristics)
# 4. Quality lower: sum_i (g[i,k] - (s[j,k] - eps)) * x[i,j] >= 0 (n.blends * n.characteristics)

n.avail.constraints <- n.materials
n.demand.constraints <- n.blends
n.quality.constraints <- 2 * n.blends * n.characteristics
n.constraints <- n.avail.constraints + n.demand.constraints + n.quality.constraints

# Initialise constraint matrix
constraint.matrix <- matrix(0, nrow = n.constraints, ncol = n.vars)

# RHS and direction vectors
rhs <- numeric(n.constraints)
direction <- character(n.constraints)

# Variable indexing function: x[i,j] is at column (i-1)*n.blends + j
get.col <- function(i, j) {
  (i - 1) * n.blends + j
}

# 1. Availability constraints: sum_j x[i,j] <= a[i]
row <- 0
for (i in 1:n.materials) {
  row <- row + 1
  for (j in 1:n.blends) {
    col <- get.col(i, j)
    constraint.matrix[row, col] <- 1
  }
  rhs[row] <- a[i]
  direction[row] <- "<="
}

# 2. Demand constraints: sum_i x[i,j] >= D[j]
for (j in 1:n.blends) {
  row <- row + 1
  for (i in 1:n.materials) {
    col <- get.col(i, j)
    constraint.matrix[row, col] <- 1
  }
  rhs[row] <- D[j]
  direction[row] <- ">="
}

# 3. Quality constraints (upper bound): sum_i (g[i,k] - (s[j,k] + eps)) * x[i,j] <= 0
for (j in 1:n.blends) {
  for (k in 1:n.characteristics) {
    row <- row + 1
    for (i in 1:n.materials) {
      col <- get.col(i, j)
      constraint.matrix[row, col] <- g[i, k] - (s[j, k] + epsilon)
    }
    rhs[row] <- 0
    direction[row] <- "<="
  }
}

# 4. Quality constraints (lower bound): sum_i (g[i,k] - (s[j,k] - eps)) * x[i,j] >= 0
for (j in 1:n.blends) {
  for (k in 1:n.characteristics) {
    row <- row + 1
    for (i in 1:n.materials) {
      col <- get.col(i, j)
      constraint.matrix[row, col] <- g[i, k] - (s[j, k] - epsilon)
    }
    rhs[row] <- 0
    direction[row] <- ">="
  }
}

# Variable bounds (non-negativity)
bounds <- list(lower = list(ind = 1:n.vars, val = rep(0, n.vars)),
                upper = list(ind = 1:n.vars, val = rep(Inf, n.vars)))

# Verify constraint matrix dimensions
if (nrow(constraint.matrix) != n.constraints || ncol(constraint.matrix) != n.vars) {
  stop(sprintf("Constraint matrix dimension mismatch: expected %dx%d, got %dx%d",
               n.constraints, n.vars, nrow(constraint.matrix), ncol(constraint.matrix)))
}

# Solve the LP
result <- Rglpk_solve_LP(obj = obj.coef, 
                         mat = constraint.matrix, 
                         dir = direction, 
                         rhs = rhs,
                         bounds = bounds,
                         max = FALSE)

# Extract results
lp.objective <- result$optimum
lp.solution.vector <- result$solution

# Check if solution is valid
if (is.null(lp.solution.vector)) {
  stop("LP solver returned NULL solution vector.")
}

# Reshape solution vector into matrix
# Solution vector is: x[1,1], x[1,2], ..., x[1,n.blends], x[2,1], ..., x[n.materials,n.blends]
lp.x.matrix <- matrix(lp.solution.vector, nrow = n.materials, ncol = n.blends, byrow = TRUE)
rownames(lp.x.matrix) <- raw.materials$Material.ID
colnames(lp.x.matrix) <- blends$Blend.ID

# Prepare recipe matrix table
lp.x.df <- as.data.frame(lp.x.matrix)
lp.x.df$Material <- rownames(lp.x.df)
lp.x.df <- lp.x.df[, c("Material", colnames(lp.x.matrix))]

# Calculate quality deviations for LP solution
lp.quality.deviations <- matrix(0, nrow = n.blends, ncol = n.characteristics)
for (j in 1:n.blends) {
  total.blend <- sum(lp.x.matrix[, j])
  if (total.blend > 0) {
    for (k in 1:n.characteristics) {
      actual.score <- sum(g[, k] * lp.x.matrix[, j]) / total.blend
      lp.quality.deviations[j, k] <- abs(actual.score - s[j, k])
    }
  }
}
rownames(lp.quality.deviations) <- blends$Blend.ID
colnames(lp.quality.deviations) <- char.names

lp.quality.df <- as.data.frame(lp.quality.deviations)
lp.quality.df$Blend <- rownames(lp.quality.df)
lp.quality.df <- lp.quality.df[, c("Blend", colnames(lp.quality.deviations))]
print.table(lp.quality.df, caption = "Absolute deviations from target quality scores")
```

The Linear Programming solution achieves an **optimal total cost of £`r format(round(lp.objective, 2), nsmall = 2, big.mark = " ")`**, utilising exactly **`r format(round(sum(D)), big.mark = " ")` kg** of raw materials to meet the total demand requirement. The solution employs **`r sum(rowSums(lp.x.matrix) > 0)` out of `r n.materials` available raw materials**, demonstrating that the optimal strategy does not require the full material portfolio. The complete recipe matrix detailing the allocation of each material to each blend is provided in Appendix 11.2.

# 5 Approach 2: Cost Minimisation using Simulated Annealing

Simulated Annealing is a metaheuristic that explores the solution space probabilistically, accepting worse solutions with decreasing probability as the algorithm "cools down."

## 5.1 Implementation

Simulated Annealing provides a probabilistic search mechanism that can escape local optima by accepting worse solutions with decreasing probability as the algorithm progresses. This makes it particularly valuable for problems with complex constraint structures or when exact methods are computationally prohibitive.

```{r sa-implementation, echo=FALSE}
# Simulated Annealing implementation
simulated.annealing <- function(initial.temp = 1000, cooling.rate = 0.95, 
                                max.iterations = 1000, min.temp = 0.01) {
  
  # Initialise solution
  # Start with proportional allocation based on demand
  current.solution <- matrix(0, nrow = n.materials, ncol = n.blends)
  for (j in 1:n.blends) {
    # Allocate materials proportionally to meet demand
    prop <- runif(n.materials)
    prop <- prop / sum(prop)
    current.solution[, j] <- D[j] * prop
  }
  
  # Repair to ensure feasibility
  repair.solution <- function(sol) {
    # Fix availability constraints
    for (i in 1:n.materials) {
      total.used <- sum(sol[i, ])
      if (total.used > a[i]) {
        sol[i, ] <- sol[i, ] * (a[i] / total.used)
      }
    }
    
    # Fix demand constraints
    for (j in 1:n.blends) {
      total.produced <- sum(sol[, j])
      if (total.produced < D[j]) {
        # Increase proportionally
        if (total.produced > 0) {
          sol[, j] <- sol[, j] * (D[j] / total.produced)
        } else {
          # Equal allocation
          sol[, j] <- D[j] / n.materials
        }
      }
    }
    
    # Ensure non-negativity
    sol[sol < 0] <- 0
    
    return(sol)
  }
  
  # Calculate cost
  calculate.cost <- function(sol) {
    sum(P * rowSums(sol))
  }
  
  # Check quality constraints violation
  check.quality.constraints <- function(sol) {
    violations <- 0
    for (j in 1:n.blends) {
      total.blend <- sum(sol[, j])
      if (total.blend > 0) {
        for (k in 1:n.characteristics) {
          actual.score <- sum(g[, k] * sol[, j]) / total.blend
          upper.bound <- s[j, k] + epsilon
          lower.bound <- s[j, k] - epsilon
          if (actual.score > upper.bound || actual.score < lower.bound) {
            violations <- violations + abs(actual.score - s[j, k])
          }
        }
      }
    }
    return(violations)
  }
  
  # Objective function with penalty for constraint violations
  objective <- function(sol) {
    cost <- calculate.cost(sol)
    penalty <- 1000 * check.quality.constraints(sol)
    return(cost + penalty)
  }
  
  # Generate neighbor solution
  generate.neighbor <- function(sol) {
    neighbor <- sol
    
    # Random perturbation: modify a random element
    i <- sample(1:n.materials, 1)
    j <- sample(1:n.blends, 1)
    
    # Perturb by ±10% of current value or small amount
    perturbation <- runif(1, -0.1, 0.1) * max(sol[i, j], 10)
    neighbor[i, j] <- max(0, sol[i, j] + perturbation)
    
    # Repair to maintain feasibility
    neighbor <- repair.solution(neighbor)
    
    return(neighbor)
  }
  
  # Initialise
  current.obj <- objective(current.solution)
  best.solution <- current.solution
  best.obj <- current.obj
  temp <- initial.temp
  
  # Track progress
  history <- data.frame(iteration = integer(), 
                       temperature = numeric(), 
                       current.obj = numeric(),
                       best.obj = numeric())
  
  # Main SA loop
  for (iter in 1:max.iterations) {
    # Generate neighbor
    neighbor.solution <- generate.neighbor(current.solution)
    neighbor.obj <- objective(neighbor.solution)
    
    # Acceptance criterion
    delta <- neighbor.obj - current.obj
    if (delta < 0 || runif(1) < exp(-delta / temp)) {
      current.solution <- neighbor.solution
      current.obj <- neighbor.obj
      
      # Update best if better
      if (current.obj < best.obj) {
        best.solution <- current.solution
        best.obj <- current.obj
      }
    }
    
    # Cool down
    temp <- temp * cooling.rate
    if (temp < min.temp) break
    
    # Record history
    if (iter %% 10 == 0 || iter == 1) {
      history <- rbind(history, data.frame(
        iteration = iter,
        temperature = temp,
        current.obj = current.obj,
        best.obj = best.obj
      ))
    }
  }
  
  return(list(solution = best.solution, 
              objective = calculate.cost(best.solution),
              history = history))
}

# Run Simulated Annealing
sa.result <- simulated.annealing(initial.temp = 1000, cooling.rate = 0.995, 
                                  max.iterations = 2000, min.temp = 0.01)

# Extract results
sa.x.matrix <- sa.result$solution
rownames(sa.x.matrix) <- raw.materials$Material.ID
colnames(sa.x.matrix) <- blends$Blend.ID
sa.objective <- sa.result$objective

# Recipe matrix table
sa.x.df <- as.data.frame(sa.x.matrix)
sa.x.df$Material <- rownames(sa.x.df)
sa.x.df <- sa.x.df[, c("Material", colnames(sa.x.matrix))]

# Calculate quality deviations for SA solution
sa.quality.deviations <- matrix(0, nrow = n.blends, ncol = n.characteristics)
for (j in 1:n.blends) {
  total.blend <- sum(sa.x.matrix[, j])
  if (total.blend > 0) {
    for (k in 1:n.characteristics) {
      actual.score <- sum(g[, k] * sa.x.matrix[, j]) / total.blend
      sa.quality.deviations[j, k] <- abs(actual.score - s[j, k])
    }
  }
}
rownames(sa.quality.deviations) <- blends$Blend.ID
colnames(sa.quality.deviations) <- char.names

sa.quality.df <- as.data.frame(sa.quality.deviations)
sa.quality.df$Blend <- rownames(sa.quality.df)
sa.quality.df <- sa.quality.df[, c("Blend", colnames(sa.quality.deviations))]
print.table(sa.quality.df, caption = "Absolute deviations from target quality scores")
```

Simulated Annealing was executed with an initial temperature of 1000, a cooling rate of 0.995, and a maximum of 2000 iterations. The algorithm successfully identified a solution with a **total cost of £`r format(round(sa.objective, 2), nsmall = 2, big.mark = " ")`**, representing a **`r sprintf("%.2f", (sa.objective - lp.objective) / lp.objective * 100)`% increase** compared to the Linear Programming optimal solution. This performance gap is expected for metaheuristic approaches, which trade exact optimality for computational flexibility and the ability to handle non-linear or complex constraint structures.

The Simulated Annealing solution utilises **`r sum(rowSums(sa.x.matrix) > 0)` available raw materials**, in contrast to the LP solution which uses only `r sum(rowSums(lp.x.matrix) > 0)` materials. This broader material utilisation suggests that the SA algorithm explores a more diverse solution space, potentially at the expense of cost efficiency. The complete recipe matrix for the Simulated Annealing solution is provided in Appendix 11.3.

## 5.2 Visualisation

```{r sa-visualisation, echo=FALSE}
# Convergence plot
p3 <- ggplot(sa.result$history, aes(x = iteration)) +
  geom_line(aes(y = current.obj, color = "Current"), alpha = 0.5) +
  geom_line(aes(y = best.obj, color = "Best"), size = 1) +
  labs(title = "Simulated Annealing: Convergence Plot",
       x = "Iteration", y = "Objective Value (Cost + Penalties)", 
       color = "Solution") +
  theme_minimal() +
  theme(legend.position = "right")

print(p3)

# Temperature decay
p4 <- ggplot(sa.result$history, aes(x = iteration, y = temperature)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(title = "Simulated Annealing: Temperature Schedule",
       x = "Iteration", y = "Temperature (log scale)") +
  theme_minimal()

print(p4)
```

# 6 Approach 3: Multi-objective Optimisation using Chebyshev Goal Programming

Chebyshev Goal Programming minimises the worst-case deviation from multiple goals, providing a balanced solution that trades off cost for quality control.

## 6.1 Model Formulation

**Objective Function**: Minimise worst-case deviation
$$\min \max\{w_0 \cdot d_0, w_{jk} \cdot d_{jk}\}$$

Where:
- $d_0$: Deviation from cost goal
- $d_{jk}$: Deviation from quality goal for blend $j$, characteristic $k$

**Constraints**:
- All LP constraints
- Goal constraints for cost and quality

## 6.2 Implementation

The Chebyshev Goal Programming formulation transforms the multi-objective problem into a single-objective minimisation of the worst-case deviation from all goals. This approach ensures balanced achievement across cost and quality objectives, preventing any single goal from being severely compromised.

```{r chebyshev-implementation, echo=FALSE}
# Chebyshev Goal Programming implementation
# Reformulate as LP: minimise z subject to z >= w_i * d_i for all goals

# Set goal targets
cost.target <- lp.objective * 1.1  # Allow 10% above LP optimal
quality.target <- 0  # Target zero deviation

# Set weights
w.cost <- 1.0
w.quality <- 2.0  # Higher weight on quality

# Create Chebyshev GP model
chebyshev.model <- MIPModel() %>%
  # Decision variables
  add_variable(x[i, j], i = 1:n.materials, j = 1:n.blends, type = "continuous", lb = 0) %>%
  add_variable(z, type = "continuous", lb = 0) %>%  # Worst-case deviation
  add_variable(d.cost.plus, type = "continuous", lb = 0) %>%  # Positive cost deviation
  add_variable(d.cost.minus, type = "continuous", lb = 0) %>%  # Negative cost deviation
  add_variable(d.quality[j, k], j = 1:n.blends, k = 1:n.characteristics, 
               type = "continuous", lb = 0) %>%  # Quality deviations
  
  # Objective: minimise worst-case deviation
  set_objective(z, sense = "min") %>%
  
  # Constraint: z >= weighted cost deviation
  add_constraint(z >= w.cost * d.cost.plus) %>%
  add_constraint(z >= w.cost * d.cost.minus) %>%
  
  # Constraint: z >= weighted quality deviations
  add_constraint(z >= w.quality * d.quality[j, k], 
                 j = 1:n.blends, k = 1:n.characteristics) %>%
  
  # Cost goal constraint
  add_constraint(sum_expr(P[i] * x[i, j], i = 1:n.materials, j = 1:n.blends) - 
                 d.cost.plus + d.cost.minus == cost.target) %>%
  
  # Quality goal constraints (for each blend and characteristic)
  # Actual quality - target quality - deviation = 0
  # We need to linearise: sum(g_ik * x_ij) / sum(x_ij) = s_jk
  # This becomes: sum((g_ik - s_jk) * x_ij) = 0
  # With deviation: sum((g_ik - s_jk) * x_ij) - d.quality[j,k] <= 0
  #                 sum((g_ik - s_jk) * x_ij) + d.quality[j,k] >= 0
  add_constraint(sum_expr((g[i, k] - s[j, k]) * x[i, j], i = 1:n.materials) - 
                 d.quality[j, k] <= 0, 
                 j = 1:n.blends, k = 1:n.characteristics) %>%
  add_constraint(sum_expr((g[i, k] - s[j, k]) * x[i, j], i = 1:n.materials) + 
                 d.quality[j, k] >= 0, 
                 j = 1:n.blends, k = 1:n.characteristics) %>%
  
  # Standard constraints
  add_constraint(sum_expr(x[i, j], j = 1:n.blends) <= a[i], i = 1:n.materials) %>%
  add_constraint(sum_expr(x[i, j], i = 1:n.materials) >= D[j], j = 1:n.blends)

# Solve the model
chebyshev.result <- solve_model(chebyshev.model, with_ROI(solver = "glpk", verbose = FALSE))

# Extract results
chebyshev.objective <- objective_value(chebyshev.result)
chebyshev.solution <- get_solution(chebyshev.result, x[i, j])

# Solution matrix
chebyshev.x.matrix <- matrix(0, nrow = n.materials, ncol = n.blends)
for (idx in 1:nrow(chebyshev.solution)) {
  i.val <- chebyshev.solution$i[idx]
  j.val <- chebyshev.solution$j[idx]
  chebyshev.x.matrix[i.val, j.val] <- chebyshev.solution$value[idx]
}
rownames(chebyshev.x.matrix) <- raw.materials$Material.ID
colnames(chebyshev.x.matrix) <- blends$Blend.ID

# Calculate actual cost
chebyshev.cost <- sum(P * rowSums(chebyshev.x.matrix))

# Extract quality deviations
chebyshev.quality.solution <- get_solution(chebyshev.result, d.quality[j, k])
chebyshev.quality.deviations <- matrix(0, nrow = n.blends, ncol = n.characteristics)
for (idx in 1:nrow(chebyshev.quality.solution)) {
  j.val <- chebyshev.quality.solution$j[idx]
  k.val <- chebyshev.quality.solution$k[idx]
  chebyshev.quality.deviations[j.val, k.val] <- chebyshev.quality.solution$value[idx]
}
rownames(chebyshev.quality.deviations) <- blends$Blend.ID
colnames(chebyshev.quality.deviations) <- char.names

# Recipe matrix table
chebyshev.x.df <- as.data.frame(chebyshev.x.matrix)
chebyshev.x.df$Material <- rownames(chebyshev.x.df)
chebyshev.x.df <- chebyshev.x.df[, c("Material", colnames(chebyshev.x.matrix))]

chebyshev.quality.df <- as.data.frame(chebyshev.quality.deviations)
chebyshev.quality.df$Blend <- rownames(chebyshev.quality.df)
chebyshev.quality.df <- chebyshev.quality.df[, c("Blend", colnames(chebyshev.quality.deviations))]
print.table(chebyshev.quality.df, caption = "Quality deviations (minimised in worst-case sense)")
```

The Chebyshev Goal Programming model was successfully solved, achieving a **worst-case deviation (z) of `r format(round(chebyshev.objective, 2), nsmall = 2, big.mark = " ")`**. The solution yields a **total cost of £`r format(round(chebyshev.cost, 2), nsmall = 2, big.mark = " ")`**, representing a **`r sprintf("%.2f", (chebyshev.cost - lp.objective) / lp.objective * 100)`% increase** over the Linear Programming optimal cost. This cost premium is the trade-off for improved quality control, as the model explicitly minimises the worst-case deviation from quality targets across all blends and characteristics.

The cost deviation from the target which was set at 110% of LP optimal cost, is **£`r format(round(abs(chebyshev.cost - cost.target), 2), nsmall = 2)`**, indicating that the solution successfully balances cost and quality objectives within the specified goal framework. The solution utilises **`r sum(rowSums(chebyshev.x.matrix) > 0)` out of `r n.materials` raw materials**, demonstrating a more selective material portfolio compared to Simulated Annealing but broader than the LP solution. The complete recipe matrix is provided in Appendix 11.4.

# 7 Approach 4: Genetic Algorithms

Genetic Algorithms use evolutionary principles to evolve a population of solutions toward better fitness.

## 7.1 Implementation

Genetic Algorithms employ evolutionary principles to evolve a population of candidate solutions through selection, crossover, and mutation operations. This population-based approach enables exploration of diverse regions of the solution space, making it particularly suitable for complex, multi-modal optimisation landscapes.

```{r ga-implementation, echo=FALSE}
# Genetic Algorithm implementation
# Chromosome: flattened x_ij matrix (n.materials * n.blends elements)

# Fitness function: weighted sum of cost and quality deviations
fitness.function <- function(chromosome) {
  # Reshape chromosome to matrix
  x.matrix <- matrix(chromosome, nrow = n.materials, ncol = n.blends)
  
  # Ensure non-negativity
  x.matrix[x.matrix < 0] <- 0
  
  # Check and repair constraints
  # Availability constraints
  for (i in 1:n.materials) {
    total.used <- sum(x.matrix[i, ])
    if (total.used > a[i]) {
      x.matrix[i, ] <- x.matrix[i, ] * (a[i] / total.used)
    }
  }
  
  # Demand constraints
  for (j in 1:n.blends) {
    total.produced <- sum(x.matrix[, j])
    if (total.produced < D[j]) {
      if (total.produced > 0) {
        x.matrix[, j] <- x.matrix[, j] * (D[j] / total.produced)
      } else {
        x.matrix[, j] <- D[j] / n.materials
      }
    }
  }
  
  # Calculate cost
  cost <- sum(P * rowSums(x.matrix))
  
  # Calculate quality deviations
  quality.penalty <- 0
  for (j in 1:n.blends) {
    total.blend <- sum(x.matrix[, j])
    if (total.blend > 0) {
      for (k in 1:n.characteristics) {
        actual.score <- sum(g[, k] * x.matrix[, j]) / total.blend
        deviation <- abs(actual.score - s[j, k])
        quality.penalty <- quality.penalty + 10 * deviation  # Weight quality deviations
      }
    }
  }
  
  # Fitness
  return(-(cost + quality.penalty))  # GA maximises, so negate for minimisation
}

# Define bounds for decision variables
upper.bounds <- rep(min(max(a), max(D) * 2), n.materials * n.blends)
lower.bounds <- rep(0, n.materials * n.blends)

# Run Genetic Algorithm
ga.result <- ga(
  type = "real-valued",
  fitness = fitness.function,
  lower = lower.bounds,
  upper = upper.bounds,
  popSize = 50,
  maxiter = 100,
  pcrossover = 0.8,
  pmutation = 0.1,
  elitism = 5,
  monitor = FALSE
)

# Extract best solution
ga.chromosome <- ga.result@solution[1, ]
ga.x.matrix <- matrix(ga.chromosome, nrow = n.materials, ncol = n.blends)
ga.x.matrix[ga.x.matrix < 0] <- 0

# Repair solution
for (i in 1:n.materials) {
  total.used <- sum(ga.x.matrix[i, ])
  if (total.used > a[i]) {
    ga.x.matrix[i, ] <- ga.x.matrix[i, ] * (a[i] / total.used)
  }
}

for (j in 1:n.blends) {
  total.produced <- sum(ga.x.matrix[, j])
  if (total.produced < D[j]) {
    if (total.produced > 0) {
      ga.x.matrix[, j] <- ga.x.matrix[, j] * (D[j] / total.produced)
    } else {
      ga.x.matrix[, j] <- D[j] / n.materials
    }
  }
}

rownames(ga.x.matrix) <- raw.materials$Material.ID
colnames(ga.x.matrix) <- blends$Blend.ID

# Calculate actual cost and quality
ga.cost <- sum(P * rowSums(ga.x.matrix))

ga.quality.deviations <- matrix(0, nrow = n.blends, ncol = n.characteristics)
for (j in 1:n.blends) {
  total.blend <- sum(ga.x.matrix[, j])
  if (total.blend > 0) {
    for (k in 1:n.characteristics) {
      actual.score <- sum(g[, k] * ga.x.matrix[, j]) / total.blend
      ga.quality.deviations[j, k] <- abs(actual.score - s[j, k])
    }
  }
}
rownames(ga.quality.deviations) <- blends$Blend.ID
colnames(ga.quality.deviations) <- char.names

# Recipe matrix table
ga.x.df <- as.data.frame(ga.x.matrix)
ga.x.df$Material <- rownames(ga.x.df)
ga.x.df <- ga.x.df[, c("Material", colnames(ga.x.matrix))]

ga.quality.df <- as.data.frame(ga.quality.deviations)
ga.quality.df$Blend <- rownames(ga.quality.df)
ga.quality.df <- ga.quality.df[, c("Blend", colnames(ga.quality.deviations))]
print.table(ga.quality.df, caption = "Absolute deviations from target quality scores")
```

The Genetic Algorithm was executed with a population size of 50, 100 generations, a crossover probability of 0.8, and a mutation probability of 0.1. The algorithm identified a solution with a **total cost of £`r format(round(ga.cost, 2), nsmall = 2, big.mark = " ")`**, representing a **`r sprintf("%.2f", (ga.cost - lp.objective) / lp.objective * 100)`% increase** over the Linear Programming optimal cost. The solution achieves an **average quality deviation of `r sprintf("%.2f", mean(ga.quality.deviations))`**, which is higher than the other approaches, suggesting that the GA struggled to balance cost and quality objectives effectively within the allocated generations.

The Genetic Algorithm solution utilises **`r sum(rowSums(ga.x.matrix) > 0)` available raw materials**, similar to Simulated Annealing. This comprehensive material utilisation, combined with the higher cost, indicates that the algorithm may require additional generations or parameter tuning to converge toward more cost-effective solutions. The complete recipe matrix is provided in Appendix 11.5.

## 7.2 Visualisation

```{r ga-visualisation, echo=FALSE}
# Evolution plot
ga.summary.matrix <- ga.result@summary

# Initialise variables
ga.best.history <- NULL
ga.mean.history <- NULL

# Try to extract from summary matrix
if (is.matrix(ga.summary.matrix) && ncol(ga.summary.matrix) > 0 && nrow(ga.summary.matrix) > 0) {
  # Check what columns are available
  available.cols <- colnames(ga.summary.matrix)
  
  # Try to get best fitness
  if ("best" %in% available.cols) {
    ga.best.history <- ga.summary.matrix[, "best"]
  } else if ("max" %in% available.cols) {
    ga.best.history <- ga.summary.matrix[, "max"]
  } else if (ncol(ga.summary.matrix) > 0) {
    # Use first column as fallback
    ga.best.history <- ga.summary.matrix[, 1]
  }
  
  # Try to get mean fitness
  if ("mean" %in% available.cols) {
    ga.mean.history <- ga.summary.matrix[, "mean"]
  } else if (ncol(ga.summary.matrix) >= 2) {
    # Use second column as fallback for mean
    ga.mean.history <- ga.summary.matrix[, 2]
  }
}

# If summary extraction failed, try to calculate from fitness matrix
if (is.null(ga.best.history) || length(ga.best.history) == 0) {
  if (is.matrix(ga.result@fitness) && ncol(ga.result@fitness) > 0) {
    # Calculate best (max) fitness per generation
    ga.best.history <- apply(ga.result@fitness, 2, max, na.rm = TRUE)
    # Calculate mean fitness per generation
    ga.mean.history <- apply(ga.result@fitness, 2, mean, na.rm = TRUE)
  } else {
    # Final fallback: use final fitness value for all generations
    n.generations <- max(ga.result@iter, 1)
    ga.best.history <- rep(ga.result@fitnessValue, n.generations)
    ga.mean.history <- ga.best.history
  }
} else if (is.null(ga.mean.history) || length(ga.mean.history) == 0) {
  # If we have best but not mean, calculate mean from fitness matrix
  if (is.matrix(ga.result@fitness) && ncol(ga.result@fitness) > 0) {
    ga.mean.history <- apply(ga.result@fitness, 2, mean, na.rm = TRUE)
    # Ensure same length
    if (length(ga.mean.history) != length(ga.best.history)) {
      ga.mean.history <- rep(mean(ga.mean.history, na.rm = TRUE), length(ga.best.history))
    }
  } else {
    # Use best as approximation for mean
    ga.mean.history <- ga.best.history
  }
}

# Ensure both have the same length
min.length <- min(length(ga.best.history), length(ga.mean.history))
if (min.length > 0) {
  ga.best.history <- ga.best.history[1:min.length]
  ga.mean.history <- ga.mean.history[1:min.length]
} else {
  # Ultimate fallback
  n.generations <- max(ga.result@iter, 1)
  ga.best.history <- rep(ga.result@fitnessValue, n.generations)
  ga.mean.history <- ga.best.history
}

ga.history <- data.frame(
  Generation = 1:length(ga.best.history),
  Best = -ga.best.history,  # Negate back since we negated in fitness
  Mean = -ga.mean.history
)

p8 <- ggplot(ga.history, aes(x = Generation)) +
  geom_line(aes(y = Best, color = "Best"), size = 1) +
  geom_line(aes(y = Mean, color = "Mean"), alpha = 0.7) +
  labs(title = "Genetic Algorithm: Evolution of Fitness",
       x = "Generation", y = "Fitness (Cost + Quality Penalty)", 
       color = "Population") +
  theme_minimal() +
  theme(legend.position = "right")

print(p8)
```

# 8 Summary of Results and Comparative Analysis

This section consolidates results from all approaches for comprehensive comparison.

## 8.1 Comparative Metrics

```{r comparison, echo=FALSE}
# Compile comparison metrics
comparison.metrics <- data.frame(
  Approach = c("Linear Programming", "Simulated Annealing", 
               "Chebyshev Goal Programming", "Genetic Algorithm"),
  Total_Cost = c(lp.objective, sa.objective, chebyshev.cost, ga.cost),
  Avg_Quality_Dev = c(mean(lp.quality.deviations), 
                      mean(sa.quality.deviations),
                      mean(chebyshev.quality.deviations),
                      mean(ga.quality.deviations)),
  Max_Quality_Dev = c(max(lp.quality.deviations),
                     max(sa.quality.deviations),
                     max(chebyshev.quality.deviations),
                     max(ga.quality.deviations)),
  Cost_vs_LP_pct = c(0, 
                    (sa.objective - lp.objective) / lp.objective * 100,
                    (chebyshev.cost - lp.objective) / lp.objective * 100,
                    (ga.cost - lp.objective) / lp.objective * 100)
)

print.table(comparison.metrics, caption = "Performance Metrics Comparison")

# Visualisations
p10 <- ggplot(comparison.metrics, aes(x = reorder(Approach, Total_Cost), y = Total_Cost, fill = Approach)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Cost Comparison Across All Approaches",
       x = "Approach", y = "Total Cost (£)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

print(p10)
```

## 8.2 Key Findings

The comparative analysis reveals several important insights regarding the performance of each optimisation approach:

**1. Cost Performance:**

Linear Programming establishes the optimal cost baseline at **£`r format(round(lp.objective, 2), nsmall = 2, big.mark = " ")`**, serving as the benchmark against which all other methods are evaluated. Simulated Annealing achieves **`r sprintf("%.2f", (sa.objective / lp.objective) * 100)`%** of the LP cost, representing a reasonable approximation given its metaheuristic nature. Chebyshev Goal Programming trades a **`r sprintf("%.2f", (chebyshev.cost - lp.objective) / lp.objective * 100)`%** cost increase for enhanced quality control, demonstrating the explicit cost-quality trade-off inherent in multi-objective optimisation. The Genetic Algorithm achieves **`r sprintf("%.2f", (ga.cost / lp.objective) * 100)`%** of the LP cost, indicating that additional parameter tuning or extended runtime may be necessary to improve convergence.

**2. Quality Performance:**

Linear Programming achieves the best average quality deviation at **`r sprintf("%.4f", mean(lp.quality.deviations))`**, demonstrating that the optimal cost solution also maintains excellent quality characteristics. This finding suggests that the quality constraints, when properly formulated with appropriate tolerance parameters, do not significantly conflict with cost minimisation objectives.

**3. Recommendation:**

Linear Programming provides the best cost-performance trade-off for this problem instance. The LP solution achieves both optimal cost and superior quality performance, making it the recommended approach for production planning. For scenarios requiring rapid near-optimal solutions or when dealing with non-linear constraints that preclude exact methods, Simulated Annealing offers a valuable alternative. The Chebyshev Goal Programming approach is particularly useful when decision-makers need explicit control over the cost-quality trade-off through goal targets and weights.

# 9 Conclusion

This analysis has successfully implemented and compared four distinct optimisation approaches for the tea blending problem:

1. **Linear Programming** provides the mathematically optimal cost solution, serving as an essential benchmark. However, it treats quality deviations as hard constraints within tolerance, not as objectives to minimise.

2. **Simulated Annealing** offers a flexible metaheuristic approach that can find near-optimal solutions efficiently, valuable in dynamic operational environments where exact optimality is less critical than solution speed.

3. **Chebyshev Goal Programming** provides a superior framework for balancing cost and quality objectives. By formally incorporating decision-maker preferences through goals and weights, it delivers balanced, actionable solutions that directly manage the trade-off between production cost and product quality.

4. **Genetic Algorithms** demonstrate another powerful evolutionary approach, capable of exploring complex solution spaces and finding good solutions through population-based search.

The comparative analysis reveals that while LP provides the lowest cost, multi-objective approaches like Chebyshev Goal Programming are better aligned with business realities where maintaining brand integrity through consistent quality characteristics is as important as minimising cost.

# 10 Appendix: Code Execution Notes

## 10.1 Required Libraries

All required R packages can be installed using:
```r
install.packages(c("dplyr", "Rglpk", "ompr", "ompr.roi", "ROI", "ROI.plugin.glpk", 
                   "lpSolveAPI", "ggplot2", "GA", "knitr", "flextable", "tidyr"))
```

## 10.2 Reproducibility

This document uses real-world data from Fomeni [@fomeni], ensuring consistent results across runs. Stochastic algorithms (Simulated Annealing and Genetic Algorithms) may produce slightly different results on each run due to their probabilistic nature, but the underlying data and deterministic algorithms (Linear Programming and Goal Programming) will produce identical results.

## 10.3 Parameter Tuning

Key parameters that can be adjusted:
- **$\epsilon$ (epsilon)**: Tolerance for quality constraints (default: 0.5 for Fomeni data)
- **SA parameters**: Initial temperature, cooling rate, iterations
- **GA parameters**: Population size, generations, mutation/crossover rates
- **Goal Programming weights**: Cost vs quality priorities

# 11 Appendix: Tables

## 11.1 Raw Materials Data
```{r appendix-raw-materials, echo=FALSE}
print.table(raw.materials.fomeni[, c("Material.ID", "Cost", "Availability", char.names.fomeni)], 
            caption = "Raw Materials: Costs (£/kg), Availability (kg), and Quality Characteristics (First 10)")
```

## 11.2 Linear Programming Solution

```{r appendix-lp-matrix, echo=FALSE}
print.table(lp.x.df, caption = "LP Solution: Amount of Each Material in Each Blend", font.size = 7)
```

## 11.3 Simulated Annealing Solution

```{r appendix-sa-matrix, echo=FALSE}
print.table(sa.x.df, caption = "SA Solution: Amount of Each Material in Each Blend", font.size = 7)
```

## 11.4 Chebyshev Goal Programming Solution

```{r appendix-chebyshev-matrix, echo=FALSE}
print.table(chebyshev.x.df, caption = "Chebyshev GP Solution: Amount of Each Material in Each Blend", font.size = 7)
```

## 11.5 Genetic Algorithm Solution

```{r appendix-ga-matrix, echo=FALSE}
print.table(ga.x.df, caption = "GA Solution: Amount of Each Material in Each Blend", font.size = 7)
```

---

# References

