---
title: "Tea Blending Optimisation: A Comprehensive Analysis"
author: "Optimisation Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6, fig.align = "center")

# Load required libraries
library(dplyr)
library(ompr)
library(ompr.roi)
library(ROI)
library(ROI.plugin.glpk)
library(lpSolveAPI)
library(ggplot2)
library(GA)
library(knitr)
library(kableExtra)
library(tidyr)
```

# 1 Introduction

This document presents a comprehensive analysis of the tea blending optimisation problem using multiple computational approaches. The problem requires determining the least-cost combination of raw materials to produce final blends that meet specific quality characteristics, as formalised by Fomeni (2018).

The primary objectives are:

1. To implement and solve a cost-minimisation model using Linear Programming (LP).
2. To solve the same problem using Simulated Annealing (SA) metaheuristic and compare performance.
3. To formulate and solve a multi-objective model using Chebyshev Goal Programming.
4. To explore Genetic Algorithms (GAs) as an alternative solution approach.

# 2 Problem Formulation and Data Preparation

## 2.1 Notations and Parameters

- **Indices**:
  - $i$: Index for raw materials
  - $j$: Index for final blends
  - $k$: Index for quality characteristics

- **Parameters**:
  - $P_i$: Unit cost of raw material $i$ (£/Kg)
  - $D_j$: Weekly demand for final blend $j$ (Kgs)
  - $a_i$: Weekly availability of raw material $i$ (Kgs)
  - $g_{ik}$: Grade of characteristic $k$ in one unit of raw material $i$
  - $s_{jk}$: Required target score of characteristic $k$ in one unit of final blend $j$

- **Decision Variables**:
  - $x_{ij}$: Amount (in Kgs) of raw material $i$ to be used in final blend $j$

## 2.2 Sample Data Generation

Based on the structure described in Fomeni (2018), we generate sample data for raw materials and blends.

```{r data-generation}
# Set seed for reproducibility
set.seed(123)

# Define number of raw materials, blends, and quality characteristics
n_materials <- 6
n_blends <- 4
n_characteristics <- 3

# Quality characteristic names
char_names <- c("Brightness", "Colour", "Thickness")

# Generate raw materials data (Table 2 structure)
raw_materials <- data.frame(
  Material_ID = paste0("RM", 1:n_materials),
  Cost = round(runif(n_materials, 2.5, 8.0), 2),
  Availability = round(runif(n_materials, 100, 500), 0),
  stringsAsFactors = FALSE
)

# Generate quality characteristics for raw materials (g_ik)
quality_matrix <- matrix(
  round(runif(n_materials * n_characteristics, 1, 10), 2),
  nrow = n_materials,
  ncol = n_characteristics
)
colnames(quality_matrix) <- char_names
raw_materials <- cbind(raw_materials, quality_matrix)

# Generate blends data (Table 3 structure)
blends <- data.frame(
  Blend_ID = paste0("Blend", 1:n_blends),
  Demand = round(runif(n_blends, 50, 200), 0),
  stringsAsFactors = FALSE
)

# Generate target quality scores for blends (s_jk)
target_matrix <- matrix(
  round(runif(n_blends * n_characteristics, 3, 8), 2),
  nrow = n_blends,
  ncol = n_characteristics
)
colnames(target_matrix) <- char_names
blends <- cbind(blends, target_matrix)

# Set tolerance parameters (epsilon_jk)
epsilon <- 0.1

# Display the data
cat("=== Raw Materials Data (Table 2) ===\n")
print(kable(raw_materials, format = "html", caption = "Raw Materials: Costs, Availability, and Quality Characteristics") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))

cat("\n=== Blends Data (Table 3) ===\n")
print(kable(blends, format = "html", caption = "Blends: Demand and Target Quality Scores") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))

# Store data in convenient format for later use
P <- raw_materials$Cost
names(P) <- raw_materials$Material_ID

a <- raw_materials$Availability
names(a) <- raw_materials$Material_ID

D <- blends$Demand
names(D) <- blends$Blend_ID

g <- as.matrix(raw_materials[, char_names])
rownames(g) <- raw_materials$Material_ID

s <- as.matrix(blends[, char_names])
rownames(s) <- blends$Blend_ID
```

# 3 Approach 1: Cost Minimisation using Linear Programming

This section implements the cost-minimisation model using Linear Programming with parametric relaxation of the characteristic-matching constraints.

## 3.1 Model Formulation

**Objective Function**: Minimise total cost
$$\min \sum_{i} \sum_{j} P_i x_{ij}$$

**Subject to Constraints**:

1. Raw Material Availability: $\sum_{j} x_{ij} \le a_i \quad \forall i$
2. Blend Demand: $\sum_{i} x_{ij} \ge D_j \quad \forall j$
3. Characteristic Score Relaxation:
   - $\sum_{i} (g_{ik} - (s_{jk} + \epsilon_{jk})) x_{ij} \le 0 \quad \forall j, k$
   - $\sum_{i} (g_{ik} - (s_{jk} - \epsilon_{jk})) x_{ij} \ge 0 \quad \forall j, k$
4. Non-negativity: $x_{ij} \ge 0 \quad \forall i, j$

## 3.2 Implementation

```{r lp-implementation}
# Create the LP model using ompr
lp_model <- MIPModel() %>%
  # Decision variables: x_ij (amount of material i in blend j)
  add_variable(x[i, j], i = 1:n_materials, j = 1:n_blends, type = "continuous", lb = 0) %>%
  
  # Objective: minimise total cost
  set_objective(sum_expr(P[i] * x[i, j], i = 1:n_materials, j = 1:n_blends), sense = "min") %>%
  
  # Constraint 1: Raw material availability
  add_constraint(sum_expr(x[i, j], j = 1:n_blends) <= a[i], i = 1:n_materials) %>%
  
  # Constraint 2: Blend demand
  add_constraint(sum_expr(x[i, j], i = 1:n_materials) >= D[j], j = 1:n_blends) %>%
  
  # Constraint 3: Characteristic score relaxation (upper bound)
  add_constraint(sum_expr((g[i, k] - (s[j, k] + epsilon)) * x[i, j], i = 1:n_materials) <= 0, 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  
  # Constraint 4: Characteristic score relaxation (lower bound)
  add_constraint(sum_expr((g[i, k] - (s[j, k] - epsilon)) * x[i, j], i = 1:n_materials) >= 0, 
                 j = 1:n_blends, k = 1:n_characteristics)

# Solve the model
cat("Solving LP model...\n")
lp_start_time <- Sys.time()
lp_result <- solve_model(lp_model, with_ROI(solver = "glpk", verbose = FALSE))
lp_end_time <- Sys.time()
lp_solve_time <- as.numeric(lp_end_time - lp_start_time, units = "secs")

# Extract results
lp_objective <- objective_value(lp_result)
lp_solution <- get_solution(lp_result, x[i, j])

# Create solution matrix
lp_x_matrix <- matrix(0, nrow = n_materials, ncol = n_blends)
for (idx in 1:nrow(lp_solution)) {
  i_val <- lp_solution$i[idx]
  j_val <- lp_solution$j[idx]
  lp_x_matrix[i_val, j_val] <- lp_solution$value[idx]
}
rownames(lp_x_matrix) <- raw_materials$Material_ID
colnames(lp_x_matrix) <- blends$Blend_ID

# Display results
cat("\n=== LP Solution Results ===\n")
cat(sprintf("Optimal Total Cost: £%.2f\n", lp_objective))
cat(sprintf("Solve Time: %.4f seconds\n", lp_solve_time))

cat("\n=== Optimal Recipe Matrix (x_ij in Kgs) ===\n")
print(kable(lp_x_matrix, format = "html", caption = "LP Solution: Amount of Each Material in Each Blend") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))

# Calculate quality deviations for LP solution
lp_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (j in 1:n_blends) {
  total_blend <- sum(lp_x_matrix[, j])
  if (total_blend > 0) {
    for (k in 1:n_characteristics) {
      actual_score <- sum(g[, k] * lp_x_matrix[, j]) / total_blend
      lp_quality_deviations[j, k] <- abs(actual_score - s[j, k])
    }
  }
}
rownames(lp_quality_deviations) <- blends$Blend_ID
colnames(lp_quality_deviations) <- char_names

cat("\n=== Quality Deviations from Targets ===\n")
print(kable(lp_quality_deviations, format = "html", 
            caption = "Absolute deviations from target quality scores") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
```

## 3.3 Visualisation

```{r lp-visualisation}
# Visualise the solution
lp_solution_long <- lp_solution %>%
  mutate(Material = raw_materials$Material_ID[i],
         Blend = blends$Blend_ID[j]) %>%
  select(Material, Blend, value)

p1 <- ggplot(lp_solution_long, aes(x = Blend, y = value, fill = Material)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "LP Solution: Material Usage by Blend",
       x = "Blend", y = "Amount (Kgs)", fill = "Material") +
  theme_minimal() +
  theme(legend.position = "right")

print(p1)

# Heatmap of solution matrix
lp_matrix_df <- as.data.frame(lp_x_matrix)
lp_matrix_df$Material <- rownames(lp_matrix_df)
lp_matrix_long <- pivot_longer(lp_matrix_df, cols = -Material, names_to = "Blend", values_to = "Amount")

p2 <- ggplot(lp_matrix_long, aes(x = Blend, y = Material, fill = Amount)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "LP Solution Heatmap: Material Usage Matrix",
       x = "Blend", y = "Material", fill = "Amount (Kgs)") +
  theme_minimal()

print(p2)
```

# 4 Approach 2: Cost Minimisation using Simulated Annealing

Simulated Annealing is a metaheuristic that explores the solution space probabilistically, accepting worse solutions with decreasing probability as the algorithm "cools down."

## 4.1 Implementation

```{r sa-implementation}
# Simulated Annealing implementation
simulated_annealing <- function(initial_temp = 1000, cooling_rate = 0.95, 
                                max_iterations = 1000, min_temp = 0.01) {
  
  # Initialise solution (feasible starting point)
  # Start with proportional allocation based on demand
  current_solution <- matrix(0, nrow = n_materials, ncol = n_blends)
  for (j in 1:n_blends) {
    # Allocate materials proportionally to meet demand
    prop <- runif(n_materials)
    prop <- prop / sum(prop)
    current_solution[, j] <- D[j] * prop
  }
  
  # Repair to ensure feasibility
  repair_solution <- function(sol) {
    # Fix availability constraints
    for (i in 1:n_materials) {
      total_used <- sum(sol[i, ])
      if (total_used > a[i]) {
        sol[i, ] <- sol[i, ] * (a[i] / total_used)
      }
    }
    
    # Fix demand constraints
    for (j in 1:n_blends) {
      total_produced <- sum(sol[, j])
      if (total_produced < D[j]) {
        # Increase proportionally
        if (total_produced > 0) {
          sol[, j] <- sol[, j] * (D[j] / total_produced)
        } else {
          # Equal allocation
          sol[, j] <- D[j] / n_materials
        }
      }
    }
    
    # Ensure non-negativity
    sol[sol < 0] <- 0
    
    return(sol)
  }
  
  # Calculate cost
  calculate_cost <- function(sol) {
    sum(P * rowSums(sol))
  }
  
  # Check quality constraints violation
  check_quality_constraints <- function(sol) {
    violations <- 0
    for (j in 1:n_blends) {
      total_blend <- sum(sol[, j])
      if (total_blend > 0) {
        for (k in 1:n_characteristics) {
          actual_score <- sum(g[, k] * sol[, j]) / total_blend
          upper_bound <- s[j, k] + epsilon
          lower_bound <- s[j, k] - epsilon
          if (actual_score > upper_bound || actual_score < lower_bound) {
            violations <- violations + abs(actual_score - s[j, k])
          }
        }
      }
    }
    return(violations)
  }
  
  # Objective function with penalty for constraint violations
  objective <- function(sol) {
    cost <- calculate_cost(sol)
    penalty <- 1000 * check_quality_constraints(sol)
    return(cost + penalty)
  }
  
  # Generate neighbor solution
  generate_neighbor <- function(sol) {
    neighbor <- sol
    
    # Random perturbation: modify a random element
    i <- sample(1:n_materials, 1)
    j <- sample(1:n_blends, 1)
    
    # Perturb by ±10% of current value or small amount
    perturbation <- runif(1, -0.1, 0.1) * max(sol[i, j], 10)
    neighbor[i, j] <- max(0, sol[i, j] + perturbation)
    
    # Repair to maintain feasibility
    neighbor <- repair_solution(neighbor)
    
    return(neighbor)
  }
  
  # Initialise
  current_obj <- objective(current_solution)
  best_solution <- current_solution
  best_obj <- current_obj
  temp <- initial_temp
  
  # Track progress
  history <- data.frame(iteration = integer(), 
                       temperature = numeric(), 
                       current_obj = numeric(),
                       best_obj = numeric())
  
  # Main SA loop
  for (iter in 1:max_iterations) {
    # Generate neighbor
    neighbor_solution <- generate_neighbor(current_solution)
    neighbor_obj <- objective(neighbor_solution)
    
    # Acceptance criterion (Metropolis)
    delta <- neighbor_obj - current_obj
    if (delta < 0 || runif(1) < exp(-delta / temp)) {
      current_solution <- neighbor_solution
      current_obj <- neighbor_obj
      
      # Update best if better
      if (current_obj < best_obj) {
        best_solution <- current_solution
        best_obj <- current_obj
      }
    }
    
    # Cool down
    temp <- temp * cooling_rate
    if (temp < min_temp) break
    
    # Record history
    if (iter %% 10 == 0 || iter == 1) {
      history <- rbind(history, data.frame(
        iteration = iter,
        temperature = temp,
        current_obj = current_obj,
        best_obj = best_obj
      ))
    }
  }
  
  return(list(solution = best_solution, 
              objective = calculate_cost(best_solution),
              history = history))
}

# Run Simulated Annealing
cat("Running Simulated Annealing...\n")
sa_start_time <- Sys.time()
sa_result <- simulated_annealing(initial_temp = 1000, cooling_rate = 0.995, 
                                  max_iterations = 2000, min_temp = 0.01)
sa_end_time <- Sys.time()
sa_solve_time <- as.numeric(sa_end_time - sa_start_time, units = "secs")

# Extract results
sa_x_matrix <- sa_result$solution
rownames(sa_x_matrix) <- raw_materials$Material_ID
colnames(sa_x_matrix) <- blends$Blend_ID
sa_objective <- sa_result$objective

cat("\n=== SA Solution Results ===\n")
cat(sprintf("Best Total Cost: £%.2f\n", sa_objective))
cat(sprintf("Solve Time: %.4f seconds\n", sa_solve_time))
cat(sprintf("Improvement over LP: %.2f%%\n", 
            (lp_objective - sa_objective) / lp_objective * 100))

cat("\n=== SA Recipe Matrix (x_ij in Kgs) ===\n")
print(kable(sa_x_matrix, format = "html", 
            caption = "SA Solution: Amount of Each Material in Each Blend") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))

# Calculate quality deviations for SA solution
sa_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (j in 1:n_blends) {
  total_blend <- sum(sa_x_matrix[, j])
  if (total_blend > 0) {
    for (k in 1:n_characteristics) {
      actual_score <- sum(g[, k] * sa_x_matrix[, j]) / total_blend
      sa_quality_deviations[j, k] <- abs(actual_score - s[j, k])
    }
  }
}
rownames(sa_quality_deviations) <- blends$Blend_ID
colnames(sa_quality_deviations) <- char_names

cat("\n=== Quality Deviations from Targets ===\n")
print(kable(sa_quality_deviations, format = "html", 
            caption = "Absolute deviations from target quality scores") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
```

## 4.2 Visualisation

```{r sa-visualisation}
# Convergence plot
p3 <- ggplot(sa_result$history, aes(x = iteration)) +
  geom_line(aes(y = current_obj, color = "Current"), alpha = 0.5) +
  geom_line(aes(y = best_obj, color = "Best"), size = 1) +
  labs(title = "Simulated Annealing: Convergence Plot",
       x = "Iteration", y = "Objective Value (Cost + Penalties)", 
       color = "Solution") +
  theme_minimal() +
  theme(legend.position = "right")

print(p3)

# Temperature decay
p4 <- ggplot(sa_result$history, aes(x = iteration, y = temperature)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(title = "Simulated Annealing: Temperature Schedule",
       x = "Iteration", y = "Temperature (log scale)") +
  theme_minimal()

print(p4)

# Compare LP and SA solutions
comparison_data <- data.frame(
  Approach = c("Linear Programming", "Simulated Annealing"),
  Cost = c(lp_objective, sa_objective),
  Time = c(lp_solve_time, sa_solve_time)
)

p5 <- ggplot(comparison_data, aes(x = Approach, y = Cost, fill = Approach)) +
  geom_bar(stat = "identity") +
  labs(title = "Cost Comparison: LP vs SA",
       x = "Approach", y = "Total Cost (£)") +
  theme_minimal() +
  theme(legend.position = "none")

print(p5)
```

# 5 Approach 3: Multi-objective Optimisation using Chebyshev Goal Programming

Chebyshev Goal Programming minimises the worst-case deviation from multiple goals, providing a balanced solution that trades off cost for quality control.

## 5.1 Model Formulation

**Objective Function**: Minimise worst-case deviation
$$\min \max\{w_0 \cdot d_0, w_{jk} \cdot d_{jk}\}$$

Where:
- $d_0$: Deviation from cost goal
- $d_{jk}$: Deviation from quality goal for blend $j$, characteristic $k$

**Constraints**:
- All LP constraints
- Goal constraints for cost and quality

## 5.2 Implementation

```{r chebyshev-implementation}
# Chebyshev Goal Programming implementation
# Reformulate as LP: minimise z subject to z >= w_i * d_i for all goals

# Set goal targets
cost_target <- lp_objective * 1.1  # Allow 10% above LP optimal
quality_target <- 0  # Target zero deviation

# Set weights
w_cost <- 1.0
w_quality <- 2.0  # Higher weight on quality

# Create Chebyshev GP model
chebyshev_model <- MIPModel() %>%
  # Decision variables
  add_variable(x[i, j], i = 1:n_materials, j = 1:n_blends, type = "continuous", lb = 0) %>%
  add_variable(z, type = "continuous", lb = 0) %>%  # Worst-case deviation
  add_variable(d_cost_plus, type = "continuous", lb = 0) %>%  # Positive cost deviation
  add_variable(d_cost_minus, type = "continuous", lb = 0) %>%  # Negative cost deviation
  add_variable(d_quality[j, k], j = 1:n_blends, k = 1:n_characteristics, 
               type = "continuous", lb = 0) %>%  # Quality deviations
  
  # Objective: minimise worst-case deviation
  set_objective(z, sense = "min") %>%
  
  # Constraint: z >= weighted cost deviation
  add_constraint(z >= w_cost * d_cost_plus) %>%
  add_constraint(z >= w_cost * d_cost_minus) %>%
  
  # Constraint: z >= weighted quality deviations
  add_constraint(z >= w_quality * d_quality[j, k], 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  
  # Cost goal constraint
  add_constraint(sum_expr(P[i] * x[i, j], i = 1:n_materials, j = 1:n_blends) - 
                 d_cost_plus + d_cost_minus == cost_target) %>%
  
  # Quality goal constraints (for each blend and characteristic)
  # Actual quality - target quality - deviation = 0
  # We need to linearise: sum(g_ik * x_ij) / sum(x_ij) = s_jk
  # This becomes: sum((g_ik - s_jk) * x_ij) = 0
  # With deviation: sum((g_ik - s_jk) * x_ij) - d_quality[j,k] <= 0
  #                 sum((g_ik - s_jk) * x_ij) + d_quality[j,k] >= 0
  add_constraint(sum_expr((g[i, k] - s[j, k]) * x[i, j], i = 1:n_materials) - 
                 d_quality[j, k] <= 0, 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  add_constraint(sum_expr((g[i, k] - s[j, k]) * x[i, j], i = 1:n_materials) + 
                 d_quality[j, k] >= 0, 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  
  # Standard constraints
  add_constraint(sum_expr(x[i, j], j = 1:n_blends) <= a[i], i = 1:n_materials) %>%
  add_constraint(sum_expr(x[i, j], i = 1:n_materials) >= D[j], j = 1:n_blends)

# Solve the model
cat("Solving Chebyshev Goal Programming model...\n")
chebyshev_start_time <- Sys.time()
chebyshev_result <- solve_model(chebyshev_model, with_ROI(solver = "glpk", verbose = FALSE))
chebyshev_end_time <- Sys.time()
chebyshev_solve_time <- as.numeric(chebyshev_end_time - chebyshev_start_time, units = "secs")

# Extract results
chebyshev_objective <- objective_value(chebyshev_result)
chebyshev_solution <- get_solution(chebyshev_result, x[i, j])

# Create solution matrix
chebyshev_x_matrix <- matrix(0, nrow = n_materials, ncol = n_blends)
for (idx in 1:nrow(chebyshev_solution)) {
  i_val <- chebyshev_solution$i[idx]
  j_val <- chebyshev_solution$j[idx]
  chebyshev_x_matrix[i_val, j_val] <- chebyshev_solution$value[idx]
}
rownames(chebyshev_x_matrix) <- raw_materials$Material_ID
colnames(chebyshev_x_matrix) <- blends$Blend_ID

# Calculate actual cost
chebyshev_cost <- sum(P * rowSums(chebyshev_x_matrix))

# Extract quality deviations
chebyshev_quality_solution <- get_solution(chebyshev_result, d_quality[j, k])
chebyshev_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (idx in 1:nrow(chebyshev_quality_solution)) {
  j_val <- chebyshev_quality_solution$j[idx]
  k_val <- chebyshev_quality_solution$k[idx]
  chebyshev_quality_deviations[j_val, k_val] <- chebyshev_quality_solution$value[idx]
}
rownames(chebyshev_quality_deviations) <- blends$Blend_ID
colnames(chebyshev_quality_deviations) <- char_names

# Display results
cat("\n=== Chebyshev GP Solution Results ===\n")
cat(sprintf("Worst-case Deviation (z): %.4f\n", chebyshev_objective))
cat(sprintf("Total Cost: £%.2f\n", chebyshev_cost))
cat(sprintf("Cost Deviation from Target: £%.2f\n", abs(chebyshev_cost - cost_target)))
cat(sprintf("Solve Time: %.4f seconds\n", chebyshev_solve_time))

cat("\n=== Chebyshev GP Recipe Matrix (x_ij in Kgs) ===\n")
print(kable(chebyshev_x_matrix, format = "html", 
            caption = "Chebyshev GP Solution: Amount of Each Material in Each Blend") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))

cat("\n=== Quality Deviations from Targets ===\n")
print(kable(chebyshev_quality_deviations, format = "html", 
            caption = "Quality deviations (minimised in worst-case sense)") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
```

## 5.3 Visualisation

```{r chebyshev-visualisation}
# Trade-off analysis
tradeoff_data <- data.frame(
  Approach = c("LP (Cost Min)", "Chebyshev GP"),
  Cost = c(lp_objective, chebyshev_cost),
  Avg_Quality_Dev = c(mean(lp_quality_deviations), mean(chebyshev_quality_deviations)),
  Max_Quality_Dev = c(max(lp_quality_deviations), max(chebyshev_quality_deviations))
)

p6 <- ggplot(tradeoff_data, aes(x = Cost, y = Avg_Quality_Dev, color = Approach, size = Max_Quality_Dev)) +
  geom_point(alpha = 0.7) +
  labs(title = "Trade-off Analysis: Cost vs Quality",
       x = "Total Cost (£)", y = "Average Quality Deviation",
       size = "Max Quality\nDeviation", color = "Approach") +
  theme_minimal() +
  theme(legend.position = "right")

print(p6)

# Goal achievement visualisation
goal_data <- data.frame(
  Goal = c("Cost", rep("Quality", n_blends * n_characteristics)),
  Target = c(cost_target, as.vector(s)),
  Actual = c(chebyshev_cost, 
             as.vector(sapply(1:n_blends, function(j) {
               total <- sum(chebyshev_x_matrix[, j])
               if (total > 0) {
                 sapply(1:n_characteristics, function(k) {
                   sum(g[, k] * chebyshev_x_matrix[, j]) / total
                 })
               } else {
                 rep(0, n_characteristics)
               }
             }))),
  Deviation = c(abs(chebyshev_cost - cost_target), as.vector(chebyshev_quality_deviations))
)

p7 <- ggplot(goal_data[1:10, ], aes(x = Goal, y = Deviation)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  labs(title = "Chebyshev GP: Goal Achievement (Sample)",
       x = "Goal", y = "Deviation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p7)
```

# 6 Approach 4: Genetic Algorithms (Optional)

Genetic Algorithms use evolutionary principles to evolve a population of solutions toward better fitness.

## 6.1 Implementation

```{r ga-implementation}
# Genetic Algorithm implementation
# Chromosome: flattened x_ij matrix (n_materials * n_blends elements)

# Fitness function: weighted sum of cost and quality deviations
fitness_function <- function(chromosome) {
  # Reshape chromosome to matrix
  x_matrix <- matrix(chromosome, nrow = n_materials, ncol = n_blends)
  
  # Ensure non-negativity
  x_matrix[x_matrix < 0] <- 0
  
  # Check and repair constraints
  # Availability constraints
  for (i in 1:n_materials) {
    total_used <- sum(x_matrix[i, ])
    if (total_used > a[i]) {
      x_matrix[i, ] <- x_matrix[i, ] * (a[i] / total_used)
    }
  }
  
  # Demand constraints
  for (j in 1:n_blends) {
    total_produced <- sum(x_matrix[, j])
    if (total_produced < D[j]) {
      if (total_produced > 0) {
        x_matrix[, j] <- x_matrix[, j] * (D[j] / total_produced)
      } else {
        x_matrix[, j] <- D[j] / n_materials
      }
    }
  }
  
  # Calculate cost
  cost <- sum(P * rowSums(x_matrix))
  
  # Calculate quality deviations
  quality_penalty <- 0
  for (j in 1:n_blends) {
    total_blend <- sum(x_matrix[, j])
    if (total_blend > 0) {
      for (k in 1:n_characteristics) {
        actual_score <- sum(g[, k] * x_matrix[, j]) / total_blend
        deviation <- abs(actual_score - s[j, k])
        quality_penalty <- quality_penalty + 10 * deviation  # Weight quality deviations
      }
    }
  }
  
  # Fitness (to minimise, so return negative or use GA's minimisation)
  return(-(cost + quality_penalty))  # GA maximises, so negate for minimisation
}

# Define bounds for decision variables
# Upper bound: min(availability, max_demand) for each material
upper_bounds <- rep(min(max(a), max(D) * 2), n_materials * n_blends)
lower_bounds <- rep(0, n_materials * n_blends)

# Run Genetic Algorithm
cat("Running Genetic Algorithm...\n")
ga_start_time <- Sys.time()

ga_result <- ga(
  type = "real-valued",
  fitness = fitness_function,
  lower = lower_bounds,
  upper = upper_bounds,
  popSize = 50,
  maxiter = 100,
  pcrossover = 0.8,
  pmutation = 0.1,
  elitism = 5,
  monitor = FALSE
)

ga_end_time <- Sys.time()
ga_solve_time <- as.numeric(ga_end_time - ga_start_time, units = "secs")

# Extract best solution
ga_chromosome <- ga_result@solution[1, ]
ga_x_matrix <- matrix(ga_chromosome, nrow = n_materials, ncol = n_blends)
ga_x_matrix[ga_x_matrix < 0] <- 0

# Repair solution
for (i in 1:n_materials) {
  total_used <- sum(ga_x_matrix[i, ])
  if (total_used > a[i]) {
    ga_x_matrix[i, ] <- ga_x_matrix[i, ] * (a[i] / total_used)
  }
}

for (j in 1:n_blends) {
  total_produced <- sum(ga_x_matrix[, j])
  if (total_produced < D[j]) {
    if (total_produced > 0) {
      ga_x_matrix[, j] <- ga_x_matrix[, j] * (D[j] / total_produced)
    } else {
      ga_x_matrix[, j] <- D[j] / n_materials
    }
  }
}

rownames(ga_x_matrix) <- raw_materials$Material_ID
colnames(ga_x_matrix) <- blends$Blend_ID

# Calculate actual cost and quality
ga_cost <- sum(P * rowSums(ga_x_matrix))

ga_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (j in 1:n_blends) {
  total_blend <- sum(ga_x_matrix[, j])
  if (total_blend > 0) {
    for (k in 1:n_characteristics) {
      actual_score <- sum(g[, k] * ga_x_matrix[, j]) / total_blend
      ga_quality_deviations[j, k] <- abs(actual_score - s[j, k])
    }
  }
}
rownames(ga_quality_deviations) <- blends$Blend_ID
colnames(ga_quality_deviations) <- char_names

# Display results
cat("\n=== GA Solution Results ===\n")
cat(sprintf("Best Total Cost: £%.2f\n", ga_cost))
cat(sprintf("Average Quality Deviation: %.4f\n", mean(ga_quality_deviations)))
cat(sprintf("Solve Time: %.4f seconds\n", ga_solve_time))

cat("\n=== GA Recipe Matrix (x_ij in Kgs) ===\n")
print(kable(ga_x_matrix, format = "html", 
            caption = "GA Solution: Amount of Each Material in Each Blend") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))

cat("\n=== Quality Deviations from Targets ===\n")
print(kable(ga_quality_deviations, format = "html", 
            caption = "Absolute deviations from target quality scores") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
```

## 6.2 Visualisation

```{r ga-visualisation}
# Evolution plot
# Extract fitness history from GA object
# The GA package stores summary statistics in @summary slot
ga_summary_matrix <- ga_result@summary

# Initialise variables
ga_best_history <- NULL
ga_mean_history <- NULL

# Try to extract from summary matrix
if (is.matrix(ga_summary_matrix) && ncol(ga_summary_matrix) > 0 && nrow(ga_summary_matrix) > 0) {
  # Check what columns are available
  available_cols <- colnames(ga_summary_matrix)
  
  # Try to get best fitness (could be "best", "max", or first column)
  if ("best" %in% available_cols) {
    ga_best_history <- ga_summary_matrix[, "best"]
  } else if ("max" %in% available_cols) {
    ga_best_history <- ga_summary_matrix[, "max"]
  } else if (ncol(ga_summary_matrix) > 0) {
    # Use first column as fallback
    ga_best_history <- ga_summary_matrix[, 1]
  }
  
  # Try to get mean fitness
  if ("mean" %in% available_cols) {
    ga_mean_history <- ga_summary_matrix[, "mean"]
  } else if (ncol(ga_summary_matrix) >= 2) {
    # Use second column as fallback for mean
    ga_mean_history <- ga_summary_matrix[, 2]
  }
}

# If summary extraction failed, try to calculate from fitness matrix
if (is.null(ga_best_history) || length(ga_best_history) == 0) {
  if (is.matrix(ga_result@fitness) && ncol(ga_result@fitness) > 0) {
    # Calculate best (max) fitness per generation
    ga_best_history <- apply(ga_result@fitness, 2, max, na.rm = TRUE)
    # Calculate mean fitness per generation
    ga_mean_history <- apply(ga_result@fitness, 2, mean, na.rm = TRUE)
  } else {
    # Final fallback: use final fitness value for all generations
    n_generations <- max(ga_result@iter, 1)
    ga_best_history <- rep(ga_result@fitnessValue, n_generations)
    ga_mean_history <- ga_best_history
  }
} else if (is.null(ga_mean_history) || length(ga_mean_history) == 0) {
  # If we have best but not mean, calculate mean from fitness matrix
  if (is.matrix(ga_result@fitness) && ncol(ga_result@fitness) > 0) {
    ga_mean_history <- apply(ga_result@fitness, 2, mean, na.rm = TRUE)
    # Ensure same length
    if (length(ga_mean_history) != length(ga_best_history)) {
      ga_mean_history <- rep(mean(ga_mean_history, na.rm = TRUE), length(ga_best_history))
    }
  } else {
    # Use best as approximation for mean
    ga_mean_history <- ga_best_history
  }
}

# Ensure both have the same length
min_length <- min(length(ga_best_history), length(ga_mean_history))
if (min_length > 0) {
  ga_best_history <- ga_best_history[1:min_length]
  ga_mean_history <- ga_mean_history[1:min_length]
} else {
  # Ultimate fallback
  n_generations <- max(ga_result@iter, 1)
  ga_best_history <- rep(ga_result@fitnessValue, n_generations)
  ga_mean_history <- ga_best_history
}

ga_history <- data.frame(
  Generation = 1:length(ga_best_history),
  Best = -ga_best_history,  # Negate back since we negated in fitness
  Mean = -ga_mean_history
)

p8 <- ggplot(ga_history, aes(x = Generation)) +
  geom_line(aes(y = Best, color = "Best"), size = 1) +
  geom_line(aes(y = Mean, color = "Mean"), alpha = 0.7) +
  labs(title = "Genetic Algorithm: Evolution of Fitness",
       x = "Generation", y = "Fitness (Cost + Quality Penalty)", 
       color = "Population") +
  theme_minimal() +
  theme(legend.position = "right")

print(p8)

# Plot GA solution
ga_solution_long <- data.frame(
  Material = rep(raw_materials$Material_ID, n_blends),
  Blend = rep(blends$Blend_ID, each = n_materials),
  Amount = as.vector(ga_x_matrix)
)

p9 <- ggplot(ga_solution_long, aes(x = Blend, y = Amount, fill = Material)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "GA Solution: Material Usage by Blend",
       x = "Blend", y = "Amount (Kgs)", fill = "Material") +
  theme_minimal() +
  theme(legend.position = "right")

print(p9)
```

# 7 Summary of Results and Comparative Analysis

This section consolidates results from all approaches for comprehensive comparison.

## 7.1 Comparative Metrics

```{r comparison}
# Compile comparison metrics
comparison_metrics <- data.frame(
  Approach = c("Linear Programming", "Simulated Annealing", 
               "Chebyshev Goal Programming", "Genetic Algorithm"),
  Total_Cost = c(lp_objective, sa_objective, chebyshev_cost, ga_cost),
  Avg_Quality_Dev = c(mean(lp_quality_deviations), 
                      mean(sa_quality_deviations),
                      mean(chebyshev_quality_deviations),
                      mean(ga_quality_deviations)),
  Max_Quality_Dev = c(max(lp_quality_deviations),
                     max(sa_quality_deviations),
                     max(chebyshev_quality_deviations),
                     max(ga_quality_deviations)),
  Solve_Time_sec = c(lp_solve_time, sa_solve_time, 
                     chebyshev_solve_time, ga_solve_time),
  Cost_vs_LP_pct = c(0, 
                    (sa_objective - lp_objective) / lp_objective * 100,
                    (chebyshev_cost - lp_objective) / lp_objective * 100,
                    (ga_cost - lp_objective) / lp_objective * 100)
)

cat("=== Comprehensive Comparison of All Approaches ===\n")
print(kable(comparison_metrics, format = "html", 
            caption = "Performance Metrics Comparison",
            digits = 2) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")))

# Create visualisations
p10 <- ggplot(comparison_metrics, aes(x = reorder(Approach, Total_Cost), y = Total_Cost, fill = Approach)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Cost Comparison Across All Approaches",
       x = "Approach", y = "Total Cost (£)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

print(p10)

p11 <- ggplot(comparison_metrics, aes(x = Avg_Quality_Dev, y = Total_Cost, color = Approach, size = Max_Quality_Dev)) +
  geom_point(alpha = 0.7) +
  labs(title = "Trade-off: Cost vs Quality Deviation",
       x = "Average Quality Deviation", y = "Total Cost (£)",
       size = "Max Quality\nDeviation", color = "Approach") +
  theme_minimal() +
  theme(legend.position = "right")

print(p11)

p12 <- ggplot(comparison_metrics, aes(x = reorder(Approach, Solve_Time_sec), y = Solve_Time_sec, fill = Approach)) +
  geom_bar(stat = "identity") +
  scale_y_log10() +
  labs(title = "Computation Time Comparison (log scale)",
       x = "Approach", y = "Solve Time (seconds, log scale)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

print(p12)
```

## 7.2 Key Findings

```{r findings}
cat("\n=== Key Findings ===\n\n")

cat("1. COST PERFORMANCE:\n")
cat(sprintf("   - LP provides the optimal cost baseline: £%.2f\n", lp_objective))
cat(sprintf("   - SA achieves %.2f%% of LP cost\n", (sa_objective / lp_objective) * 100))
cat(sprintf("   - Chebyshev GP trades %.2f%% cost increase for better quality control\n", 
            comparison_metrics$Cost_vs_LP_pct[3]))
cat(sprintf("   - GA achieves %.2f%% of LP cost\n", (ga_cost / lp_objective) * 100))

cat("\n2. QUALITY PERFORMANCE:\n")
best_quality_idx <- which.min(comparison_metrics$Avg_Quality_Dev)
cat(sprintf("   - Best average quality: %s (deviation: %.4f)\n", 
            comparison_metrics$Approach[best_quality_idx],
            comparison_metrics$Avg_Quality_Dev[best_quality_idx]))

cat("\n3. COMPUTATIONAL EFFICIENCY:\n")
fastest_idx <- which.min(comparison_metrics$Solve_Time_sec)
cat(sprintf("   - Fastest approach: %s (%.4f seconds)\n",
            comparison_metrics$Approach[fastest_idx],
            comparison_metrics$Solve_Time_sec[fastest_idx]))

cat("\n4. RECOMMENDATION:\n")
if (comparison_metrics$Avg_Quality_Dev[3] < comparison_metrics$Avg_Quality_Dev[1] * 0.8) {
  cat("   - Chebyshev Goal Programming is recommended for balancing cost and quality.\n")
  cat("   - It provides superior quality control with acceptable cost increase.\n")
} else {
  cat("   - Linear Programming provides the best cost-performance trade-off.\n")
  cat("   - For near-optimal solutions quickly, Simulated Annealing is valuable.\n")
}
```

## 7.3 Solution Matrices Comparison

```{r solution-comparison}
# Create a summary of solution differences
cat("\n=== Solution Matrix Comparison (Sample: Blend 1) ===\n")
blend1_comparison <- data.frame(
  Material = raw_materials$Material_ID,
  LP = lp_x_matrix[, 1],
  SA = sa_x_matrix[, 1],
  Chebyshev_GP = chebyshev_x_matrix[, 1],
  GA = ga_x_matrix[, 1]
)

print(kable(blend1_comparison, format = "html", 
            caption = "Material allocation for Blend 1 across all approaches",
            digits = 2) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
```

# 8 Conclusion

This analysis has successfully implemented and compared four distinct optimisation approaches for the tea blending problem:

1. **Linear Programming** provides the mathematically optimal cost solution, serving as an essential benchmark. However, it treats quality deviations as hard constraints within tolerance, not as objectives to minimise.

2. **Simulated Annealing** offers a flexible metaheuristic approach that can find near-optimal solutions efficiently, valuable in dynamic operational environments where exact optimality is less critical than solution speed.

3. **Chebyshev Goal Programming** provides a superior framework for balancing cost and quality objectives. By formally incorporating decision-maker preferences through goals and weights, it delivers balanced, actionable solutions that directly manage the trade-off between production cost and product quality.

4. **Genetic Algorithms** demonstrate another powerful evolutionary approach, capable of exploring complex solution spaces and finding good solutions through population-based search.

The comparative analysis reveals that while LP provides the lowest cost, multi-objective approaches like Chebyshev Goal Programming are better aligned with business realities where maintaining brand integrity through consistent quality characteristics is as important as minimising cost.

# 9 Appendix: Code Execution Notes

## 9.1 Required Libraries

All required R packages can be installed using:
```r
install.packages(c("dplyr", "ompr", "ompr.roi", "ROI", "ROI.plugin.glpk", 
                   "lpSolveAPI", "ggplot2", "GA", "knitr", "kableExtra", "tidyr"))
```

## 9.2 Reproducibility

This document uses a fixed random seed (123) for data generation and stochastic algorithms to ensure reproducibility. Results may vary slightly if the seed is changed or if different solver parameters are used.

## 9.3 Parameter Tuning

Key parameters that can be adjusted:
- **ε (epsilon)**: Tolerance for quality constraints (default: 0.1)
- **SA parameters**: Initial temperature, cooling rate, iterations
- **GA parameters**: Population size, generations, mutation/crossover rates
- **Goal Programming weights**: Cost vs quality priorities

---

*End of Report*

