---
title: "Tea Blending Optimisation: A Comprehensive Analysis"
subtitle: "STA5074Z Prescriptive Analytics"
author: "Tinotenda Muzambi - MZMTIN002"
date: "`r Sys.Date()`"
bibliography: references.bib
csl: apa-numeric-superscript-brackets.csl
output:
  pdf_document:
    toc: true
    latex_engine: xelatex  # Support Unicode characters
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 5, fig.height = 3, fig.align = "center")

# Load required libraries
library(dplyr)
library(Rglpk)
library(ompr)
library(ompr.roi)
library(ROI)
library(ROI.plugin.glpk)
library(lpSolveAPI)
library(ggplot2)
library(GA)
library(knitr)
library(flextable) 
library(tidyr)

# Helper function to print tables 
print_table <- function(data, caption = "") {
  ft <- flextable(data) %>%
    set_caption(caption) %>%
    theme_zebra(odd_header = "#CFCFCF", odd_body = "#EFEFEF") %>%
    autofit()
  
  if (knitr::is_latex_output()) {
    ft <- ft %>%
      fontsize(size = 9) %>%
      set_table_properties(layout = "autofit")
  }
  
  return(ft)
}
```

# 1 Introduction
Blending problems are a classic challenge in operations research, arising in diverse fields such as food processing, metallurgy, and energy. In the tea industry, blending is crucial to achieve consistent product quality, balance raw material costs, and adapt to fluctuating supplies. Every cup of tea on the supermarket shelf is the result of a carefully orchestrated process. Different grades and types of tea leaves from various origins are combined to meet precise targets for taste, colour, aroma and other quality parameters, all while minimising production costs and ensuring supply constraints are not violated.

In the context of tea manufacturing, the blending optimisation problem involves selecting the optimal composition of raw teas to satisfy desired blend qualities for multiple final blends. This not only impacts the sensory profile and brand identity of each product, but also has substantial operational and economic implications. By applying advanced computational techniques, such as mathematical programming and metaheuristics, producers can automate and enhance the blending decision-making process, yielding better margins, consistent quality and greater responsiveness to market demands.

Below, we formally define the tea blending optimisation problem and systematically explore several approaches for solving it. The problem requires determining the least-cost combination of raw materials to produce final blends that meet specific quality characteristics, as formalised by Fomeni [@fomeni].

The primary objectives are:

1. To implement and solve a cost-minimisation model using Linear Programming (LP).
2. To solve the same problem using Simulated Annealing (SA) metaheuristic and compare performance.
3. To formulate and solve a multi-objective model using Chebyshev Goal Programming.
4. To explore Genetic Algorithms (GAs) as an alternative solution approach.

# 2 Problem Formulation and Data Preparation

## 2.1 Notations and Parameters

- **Indices**:
  - $i$: Index for raw materials
  - $j$: Index for final blends
  - $k$: Index for quality characteristics

- **Parameters**:
  - $P_i$: Unit cost of raw material $i$ (£/Kg)
  - $D_j$: Weekly demand for final blend $j$ (Kgs)
  - $a_i$: Weekly availability of raw material $i$ (Kgs)
  - $g_{ik}$: Grade of characteristic $k$ in one unit of raw material $i$
  - $s_{jk}$: Required target score of characteristic $k$ in one unit of final blend $j$

- **Decision Variables**:
  - $x_{ij}$: Amount (in Kgs) of raw material $i$ to be used in final blend $j$

## 2.2 Data Source

This project uses the real-world data from Fomeni [@fomeni], extracted from tables 2 and 3 of the paper. These data represent actual tea blending scenarios from a UK-based tea company. The data are loaded and prepared in section 3 below.

# 3 Data Implementation (Fomeni 2018)

This section implements the tea blending optimisation problem using the actual data provided in Fomeni [@fomeni]. The data consist of 60 raw materials and 10 final blends, each with 6 quality characteristics.

## 3.1 Data Loading

The dataset comprises **60 raw materials** and **10 final blends**, each characterised by **6 quality attributes** (C1 through C6). Raw material prices range from £0.0120 to £0.0470 per kilogram, with a mean price of £0.0217 per kilogram. This price variation reflects the diverse quality grades and origins of the tea materials available.

```{r fomeni-data, echo=FALSE}
# Real data from Fomeni (2018) - Table 2: Raw Materials
# Note: Prices are in £0.01/kg, so we convert to £/kg by dividing by 100
# Quality characteristics: C1, C2, C3, C4, C5, C6

raw_materials_fomeni <- data.frame(
  Material_ID = paste0("R", 1:60),
  C1 = c(5.0, 5.0, 5.0, 5.0, 5.0, 7.0, 7.0, 8.0, 8.0, 8.0, 2.0, 2.0, 2.0, 3.0, 4.0, 4.0, 7.0, 4.0, 4.0, 5.0, 5.0, 5.0, 5.0, 6.0, 6.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 9.0, 2.0, 5.0, 5.0, 5.0, 5.0, 2.0, 4.0, 5.0, 5.0, 2.0, 5.0, 5.0, 5.0, 6.0, 6.0, 7.0, 8.0, 8.0, 5.0, 5.0, 5.0, 7.0, 8.0, 8.0, 2.0, 4.0, 6.0, 6.0),
  C2 = c(4.0, 3.0, 4.0, 3.0, 4.0, 4.0, 4.0, 4.0, 4.0, 4.0, 2.0, 2.0, 0.5, 0.5, 0.5, 2.0, 0.5, 8.0, 8.0, 6.0, 6.0, 6.0, 6.0, 6.0, 6.0, 5.0, 5.0, 5.0, 5.0, 5.0, 5.0, 5.0, 0.5, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 3.0, 3.0, 3.0, 3.0, 3.0, 2.0, 2.0, 0.5, 1.0, 1.0, 1.0, 2.0, 8.0, 5.0, 5.0),
  C3 = c(9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 9.0, 3.0, 3.0, 1.0, 1.0, 1.0, 3.0, 1.0, 5.0, 5.0, 4.0, 4.0, 4.0, 4.0, 4.0, 4.0, 4.0, 6.0, 6.0, 7.0, 6.0, 6.0, 6.0, 1.0, 2.0, 2.0, 2.0, 3.0, 3.0, 3.0, 2.0, 3.0, 5.0, 4.0, 4.0, 4.0, 4.0, 4.0, 3.0, 3.0, 3.0, 2.0, 2.0, 1.0, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0),
  C4 = c(6.0, 4.0, 6.0, 4.0, 6.0, 6.0, 6.0, 6.0, 6.0, 6.0, 2.0, 2.0, 5.0, 5.0, 5.0, 2.0, 5.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 5.0, 3.0, 3.0, 3.0, 3.0, 2.0, 2.0, 3.0, 3.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 3.0, 3.0, 3.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0, 2.0),
  C5 = c(5.0, 5.0, 5.0, 5.0, 5.0, 7.0, 7.0, 8.0, 8.0, 8.0, 2.0, 2.0, 2.0, 3.0, 4.0, 4.0, 7.0, 4.0, 4.0, 5.0, 5.0, 5.0, 5.0, 6.0, 6.0, 8.0, 8.0, 8.0, 8.0, 8.0, 8.0, 9.0, 2.0, 5.0, 5.0, 5.0, 5.0, 2.0, 4.0, 5.0, 5.0, 2.0, 5.0, 5.0, 5.0, 6.0, 6.0, 7.0, 8.0, 8.0, 5.0, 5.0, 5.0, 7.0, 8.0, 8.0, 2.0, 4.0, 6.0, 6.0),
  C6 = c(5.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 5.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 9.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 5.0, 1.0, 5.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 5.0, 1.0, 5.0, 1.0, 5.0, 5.0, 1.0, 5.0, 1.0, 1.0, 5.0, 5.0, 1.0, 1.0, 1.0, 1.0, 5.0),
  Price_001_per_kg = c(2.40, 1.67, 2.35, 1.70, 2.30, 2.65, 2.65, 2.94, 2.84, 2.84, 1.30, 1.30, 1.20, 1.70, 1.50, 2.50, 1.80, 1.30, 1.30, 1.80, 1.90, 1.80, 1.50, 1.90, 1.80, 2.60, 2.00, 2.00, 2.10, 2.00, 1.70, 2.40, 1.40, 1.60, 1.60, 1.60, 1.60, 1.30, 2.50, 1.60, 1.70, 2.10, 2.40, 2.50, 2.50, 2.70, 2.80, 3.30, 3.40, 3.60, 1.80, 1.70, 2.10, 2.90, 4.70, 4.20, 1.30, 1.60, 1.70, 4.00),
  Availability_kg = c(5000, 1000, 25000, 1000, 25000, 20000, 20000, 1000, 15000, 15000, 40000, 80000, 60000, 5000, 5000, 20000, 20000, 20000, 20000, 40000, 40000, 20000, 5000, 1000, 500, 1000, 40000, 40000, 25000, 30000, 5000, 5000, 10000, 10000, 10000, 10000, 10000, 20000, 15000, 2000, 5000, 1000, 5000, 5000, 5000, 2000, 2000, 1000, 1000, 1000, 1000, 5000, 5000, 1000, 1000, 1000, 5000, 40000, 5000, 1000),
  stringsAsFactors = FALSE
)

# Convert price from £0.01/kg to £/kg
raw_materials_fomeni$Cost <- raw_materials_fomeni$Price_001_per_kg / 100
raw_materials_fomeni$Availability <- raw_materials_fomeni$Availability_kg

# Extract quality characteristics matrix
char_names_fomeni <- c("C1", "C2", "C3", "C4", "C5", "C6")
quality_matrix_fomeni <- as.matrix(raw_materials_fomeni[, char_names_fomeni])

# Table 3: Blends data
blends_fomeni <- data.frame(
  Blend_ID = paste0("B", 1:10),
  C1 = c(2.30, 3.80, 6.60, 6.20, 4.70, 8.40, 4.00, 5.90, 7.00, 4.00),
  C2 = c(3.00, 4.70, 5.20, 5.20, 5.10, 4.70, 0.50, 0.50, 3.90, 2.00),
  C3 = c(3.00, 2.90, 6.10, 5.30, 4.50, 6.70, 1.00, 1.00, 9.00, 3.00),
  C4 = c(2.00, 3.00, 6.90, 6.40, 5.00, 7.30, 5.00, 3.90, 5.90, 2.00),
  C5 = c(2.32, 3.80, 6.60, 6.20, 4.70, 8.40, 4.00, 5.90, 7.00, 4.00),
  C6 = c(1.00, 1.00, 1.00, 1.00, 1.00, 1.00, 1.00, 1.00, 1.00, 9.00),
  Demand_kg = c(100000, 100000, 60000, 100000, 100000, 50000, 10000, 5000, 10000, 20000),
  stringsAsFactors = FALSE
)

blends_fomeni$Demand <- blends_fomeni$Demand_kg

# Extract target quality matrix
target_matrix_fomeni <- as.matrix(blends_fomeni[, char_names_fomeni])

# Display first few rows of raw materials
print_table(raw_materials_fomeni[1:10, c("Material_ID", "Cost", "Availability", char_names_fomeni)], 
            caption = "Raw Materials: Costs (£/kg), Availability (kg), and Quality Characteristics (First 10)")

# Display all blends
print_table(blends_fomeni[, c("Blend_ID", "Demand", char_names_fomeni)], 
            caption = "Blends: Demand (kg) and Target Quality Scores")

# Prepare data vectors for optimisation models
n_materials_fomeni <- nrow(raw_materials_fomeni)
n_blends_fomeni <- nrow(blends_fomeni)
n_characteristics_fomeni <- length(char_names_fomeni)

# Cost vector P[i]
P_fomeni <- raw_materials_fomeni$Cost

# Availability vector a[i]
a_fomeni <- raw_materials_fomeni$Availability

# Demand vector D[j]
D_fomeni <- blends_fomeni$Demand

# Quality matrices
g_fomeni <- quality_matrix_fomeni  # g[i,k]
s_fomeni <- target_matrix_fomeni   # s[j,k]

# Set tolerance parameter (epsilon) - may need adjustment for real data
epsilon_fomeni <- 0.5  # Start with a reasonable tolerance

# Feasibility check
total_availability_fomeni <- sum(a_fomeni)
total_demand_fomeni <- sum(D_fomeni)

# Create standard variable names for use in optimisation models
# (These will be used by all subsequent optimisation approaches)
n_materials <- n_materials_fomeni
n_blends <- n_blends_fomeni
n_characteristics <- n_characteristics_fomeni
char_names <- char_names_fomeni
P <- P_fomeni
a <- a_fomeni
D <- D_fomeni
g <- g_fomeni
s <- s_fomeni
epsilon <- epsilon_fomeni
raw_materials <- raw_materials_fomeni
blends <- blends_fomeni

# Create named versions for display
P_named <- P_fomeni
names(P_named) <- raw_materials_fomeni$Material_ID
a_named <- a_fomeni
names(a_named) <- raw_materials_fomeni$Material_ID
D_named <- D_fomeni
names(D_named) <- blends_fomeni$Blend_ID
g_named <- g_fomeni
rownames(g_named) <- raw_materials_fomeni$Material_ID
s_named <- s_fomeni
rownames(s_named) <- blends_fomeni$Blend_ID
```

A feasibility analysis reveals that the total raw material availability is **830,500 kg**, while total demand across all blends is **555,000 kg**. This represents a **49.64% surplus** in availability, indicating that supply constraints are not binding and the problem has sufficient flexibility for optimisation. The surplus availability provides the solver with multiple feasible solution paths, which is advantageous for finding cost-optimal allocations.

## 3.2 Data Characteristics

The following analysis provides a comprehensive overview of the dataset's key characteristics, enabling a deeper understanding of the problem structure before optimisation.

```{r fomeni-data-analysis, echo=FALSE}
# Analyse the data characteristics and create summary tables and visualisations

# Create summary statistics table for raw materials
price_stats <- data.frame(
  Statistic = c("Mean", "Median", "Minimum", "Maximum", "Standard Deviation"),
  Value = c(
    sprintf("£%.4f", mean(P_fomeni)),
    sprintf("£%.4f", median(P_fomeni)),
    sprintf("£%.4f", min(P_fomeni)),
    sprintf("£%.4f", max(P_fomeni)),
    sprintf("£%.4f", sd(P_fomeni))
  )
)

availability_stats <- data.frame(
  Statistic = c("Mean", "Median", "Minimum", "Maximum", "Total"),
  Value_kg = c(
    sprintf("%.0f", mean(a_fomeni)),
    sprintf("%.0f", median(a_fomeni)),
    sprintf("%.0f", min(a_fomeni)),
    sprintf("%.0f", max(a_fomeni)),
    sprintf("%.0f", sum(a_fomeni))
  )
)

# Quality characteristics summary
quality_summary <- data.frame(
  Characteristic = char_names_fomeni,
  Minimum = apply(g_fomeni, 2, min),
  Maximum = apply(g_fomeni, 2, max),
  Mean = apply(g_fomeni, 2, mean),
  Median = apply(g_fomeni, 2, median),
  Std_Dev = apply(g_fomeni, 2, sd)
)

# Blend demand summary
demand_stats <- data.frame(
  Statistic = c("Mean", "Median", "Minimum", "Maximum", "Total"),
  Value_kg = c(
    sprintf("%.0f", mean(D_fomeni)),
    sprintf("%.0f", median(D_fomeni)),
    sprintf("%.0f", min(D_fomeni)),
    sprintf("%.0f", max(D_fomeni)),
    sprintf("%.0f", sum(D_fomeni))
  )
)

# Display summary tables
print_table(price_stats, caption = "Price Distribution Statistics")

print_table(availability_stats, caption = "Availability Distribution Statistics (kg)")

print_table(quality_summary, caption = "Quality Characteristics Summary Statistics")

print_table(demand_stats, caption = "Blend Demand Statistics (kg)")
```

```{r fomeni-data-visualisations, echo=FALSE}
# Create visualisations of data characteristics

# 1. Price distribution histogram
p_price <- ggplot(data.frame(Price = P_fomeni), aes(x = Price)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, colour = "white") +
  geom_vline(aes(xintercept = mean(Price)), linetype = "dashed", colour = "red", size = 1) +
  labs(title = "Distribution of Raw Material Prices",
       x = "Price (£/kg)", y = "Frequency") +
  theme_minimal()

print(p_price)

# 2. Availability distribution (log scale due to wide range)
p_avail <- ggplot(data.frame(Availability = a_fomeni), aes(x = Availability)) +
  geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.7, colour = "white") +
  scale_x_log10(labels = scales::comma) +
  geom_vline(aes(xintercept = median(Availability)), linetype = "dashed", colour = "red", size = 1) +
  labs(title = "Distribution of Raw Material Availability",
       x = "Availability (kg, log scale)", y = "Frequency") +
  theme_minimal()

print(p_avail)

# 3. Quality characteristics boxplot
quality_long <- data.frame(
  Characteristic = rep(char_names_fomeni, each = n_materials_fomeni),
  Value = as.vector(g_fomeni)
)

p_quality <- ggplot(quality_long, aes(x = Characteristic, y = Value, fill = Characteristic)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
  labs(title = "Distribution of Quality Characteristics Across Raw Materials",
       x = "Quality Characteristic", y = "Score") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

print(p_quality)

# 4. Blend demand bar chart
blend_demand_df <- data.frame(
  Blend = blends_fomeni$Blend_ID,
  Demand = D_fomeni
)

p_demand <- ggplot(blend_demand_df, aes(x = reorder(Blend, Demand), y = Demand, fill = Demand)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Demand by Blend",
       x = "Blend", y = "Demand (kg)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip()

print(p_demand)

# 5. Price vs Availability scatter plot
price_avail_df <- data.frame(
  Price = P_fomeni,
  Availability = a_fomeni,
  Material = raw_materials_fomeni$Material_ID
)

p_price_avail <- ggplot(price_avail_df, aes(x = Availability, y = Price)) +
  geom_point(alpha = 0.6, size = 2, colour = "steelblue") +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "£")) +
  labs(title = "Price vs Availability of Raw Materials",
       x = "Availability (kg, log scale)", y = "Price (£/kg)") +
  theme_minimal()

print(p_price_avail)

# 6. Quality characteristics heatmap (sample of materials)
# Show first 20 materials for readability
sample_materials <- 1:min(20, n_materials_fomeni)
quality_sample <- g_fomeni[sample_materials, ]
rownames(quality_sample) <- raw_materials_fomeni$Material_ID[sample_materials]

# Create proper ordering for materials (R1, R2, R3, ..., R20 instead of R1, R10, R11, ...)
sample_material_order <- raw_materials_fomeni$Material_ID[sample_materials][
  order(as.numeric(sub("R", "", raw_materials_fomeni$Material_ID[sample_materials])))
]

# Convert to long format for heatmap
quality_melt <- data.frame(
  Material = rep(rownames(quality_sample), ncol(quality_sample)),
  Characteristic = rep(colnames(quality_sample), each = nrow(quality_sample)),
  Value = as.vector(quality_sample)
)
# Convert Material to factor with proper ordering
quality_melt$Material <- factor(quality_melt$Material, levels = sample_material_order)

p_heatmap <- ggplot(quality_melt, aes(x = Characteristic, y = Material, fill = Value)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 5) +
  labs(title = "Quality Characteristics Heatmap (First 20 Materials)",
       x = "Quality Characteristic", y = "Raw Material", fill = "Score") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8)) +
  scale_y_discrete(limits = rev(sample_material_order))  # Reverse so R1 is at top

print(p_heatmap)
```

# 4 Approach 1: Cost Minimisation using Linear Programming

This section implements the cost-minimisation model using Linear Programming with parametric relaxation of the characteristic-matching constraints. This involves implementing a cost-minimisation LP model that finds the optimal combination of raw materials to produce blends meeting demand and quality constraints. This implementation uses Rglpk which provides a straightforward matrix-based formulation that is well-suited for this problem type.

## 4.1 Model Formulation

**Objective Function**: Minimise total cost
$$\min \sum_{i} \sum_{j} P_i x_{ij}$$

**Subject to Constraints**:

1. Raw Material Availability: $\sum_{j} x_{ij} \le a_i \quad \forall i$
2. Blend Demand: $\sum_{i} x_{ij} \ge D_j \quad \forall j$
3. Characteristic Score Relaxation:
   - $\sum_{i} (g_{ik} - (s_{jk} + \epsilon_{jk})) x_{ij} \le 0 \quad \forall j, k$
   - $\sum_{i} (g_{ik} - (s_{jk} - \epsilon_{jk})) x_{ij} \ge 0 \quad \forall j, k$
4. Non-negativity: $x_{ij} \ge 0 \quad \forall i, j$

## 4.2 Implementation

The Linear Programming model was successfully solved using the Rglpk solver, which employs the GNU Linear Programming Kit (GLPK) algorithm. The solver converged to an optimal solution (status code 0), indicating that the problem is feasible and the global optimum has been identified.

```{r lp-implementation, echo=FALSE}
# Implementation using Rglpk

# Number of decision variables: n_materials * n_blends (x[i,j] for all i,j)
n_vars <- n_materials * n_blends

# Objective function: minimise cost
# Objective coefficients: P[i] for each x[i,j]
# Variable ordering: x[1,1], x[1,2], ..., x[1,n_blends], x[2,1], ..., x[n_materials,n_blends]
obj_coef <- rep(P, n_blends)  # Repeat P for each blend

# Build constraint matrix
# Constraints in order:
# 1. Availability: sum_j x[i,j] <= a[i] for each i (n_materials constraints)
# 2. Demand: sum_i x[i,j] >= D[j] for each j (n_blends constraints)  
# 3. Quality upper: sum_i (g[i,k] - (s[j,k] + eps)) * x[i,j] <= 0 (n_blends * n_characteristics)
# 4. Quality lower: sum_i (g[i,k] - (s[j,k] - eps)) * x[i,j] >= 0 (n_blends * n_characteristics)

n_avail_constraints <- n_materials
n_demand_constraints <- n_blends
n_quality_constraints <- 2 * n_blends * n_characteristics
n_constraints <- n_avail_constraints + n_demand_constraints + n_quality_constraints

# Initialize constraint matrix
constraint_matrix <- matrix(0, nrow = n_constraints, ncol = n_vars)

# RHS and direction vectors
rhs <- numeric(n_constraints)
direction <- character(n_constraints)

# Variable indexing function: x[i,j] is at column (i-1)*n_blends + j
get_col <- function(i, j) {
  (i - 1) * n_blends + j
}

# 1. Availability constraints: sum_j x[i,j] <= a[i]
row <- 0
for (i in 1:n_materials) {
  row <- row + 1
  for (j in 1:n_blends) {
    col <- get_col(i, j)
    constraint_matrix[row, col] <- 1
  }
  rhs[row] <- a[i]
  direction[row] <- "<="
}

# 2. Demand constraints: sum_i x[i,j] >= D[j]
for (j in 1:n_blends) {
  row <- row + 1
  for (i in 1:n_materials) {
    col <- get_col(i, j)
    constraint_matrix[row, col] <- 1
  }
  rhs[row] <- D[j]
  direction[row] <- ">="
}

# 3. Quality constraints (upper bound): sum_i (g[i,k] - (s[j,k] + eps)) * x[i,j] <= 0
for (j in 1:n_blends) {
  for (k in 1:n_characteristics) {
    row <- row + 1
    for (i in 1:n_materials) {
      col <- get_col(i, j)
      constraint_matrix[row, col] <- g[i, k] - (s[j, k] + epsilon)
    }
    rhs[row] <- 0
    direction[row] <- "<="
  }
}

# 4. Quality constraints (lower bound): sum_i (g[i,k] - (s[j,k] - eps)) * x[i,j] >= 0
for (j in 1:n_blends) {
  for (k in 1:n_characteristics) {
    row <- row + 1
    for (i in 1:n_materials) {
      col <- get_col(i, j)
      constraint_matrix[row, col] <- g[i, k] - (s[j, k] - epsilon)
    }
    rhs[row] <- 0
    direction[row] <- ">="
  }
}

# Variable bounds (non-negativity)
bounds <- list(lower = list(ind = 1:n_vars, val = rep(0, n_vars)),
                upper = list(ind = 1:n_vars, val = rep(Inf, n_vars)))

# Verify constraint matrix dimensions
if (nrow(constraint_matrix) != n_constraints || ncol(constraint_matrix) != n_vars) {
  stop(sprintf("Constraint matrix dimension mismatch: expected %dx%d, got %dx%d",
               n_constraints, n_vars, nrow(constraint_matrix), ncol(constraint_matrix)))
}

# Solve the LP
result <- Rglpk_solve_LP(obj = obj_coef, 
                         mat = constraint_matrix, 
                         dir = direction, 
                         rhs = rhs,
                         bounds = bounds,
                         max = FALSE)

# Extract results
lp_objective <- result$optimum
lp_solution_vector <- result$solution

# Check if solution is valid
if (is.null(lp_solution_vector)) {
  stop("LP solver returned NULL solution vector.")
}

# Reshape solution vector into matrix
# Solution vector is: x[1,1], x[1,2], ..., x[1,n_blends], x[2,1], ..., x[n_materials,n_blends]
lp_x_matrix <- matrix(lp_solution_vector, nrow = n_materials, ncol = n_blends, byrow = TRUE)
rownames(lp_x_matrix) <- raw_materials$Material_ID
colnames(lp_x_matrix) <- blends$Blend_ID

# Prepare recipe matrix table (full table moved to Appendix)
lp_x_df <- as.data.frame(lp_x_matrix)
lp_x_df$Material <- rownames(lp_x_df)
lp_x_df <- lp_x_df[, c("Material", colnames(lp_x_matrix))]

# Calculate quality deviations for LP solution
lp_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (j in 1:n_blends) {
  total_blend <- sum(lp_x_matrix[, j])
  if (total_blend > 0) {
    for (k in 1:n_characteristics) {
      actual_score <- sum(g[, k] * lp_x_matrix[, j]) / total_blend
      lp_quality_deviations[j, k] <- abs(actual_score - s[j, k])
    }
  }
}
rownames(lp_quality_deviations) <- blends$Blend_ID
colnames(lp_quality_deviations) <- char_names

lp_quality_df <- as.data.frame(lp_quality_deviations)
lp_quality_df$Blend <- rownames(lp_quality_df)
lp_quality_df <- lp_quality_df[, c("Blend", colnames(lp_quality_deviations))]
print_table(lp_quality_df, caption = "Absolute deviations from target quality scores")
```

The Linear Programming solution achieves an **optimal total cost of £9,087.69**, utilising exactly **555,000 kg** of raw materials to meet the total demand requirement. The solution employs **52 out of 60 available raw materials**, demonstrating that the optimal strategy does not require the full material portfolio. The materials utilised in the optimal solution include R1, R2, R3, R4, R6, R7, R8, R9, R10, R11, R12, R13, R14, R15, R16, R17, R18, R19, R20, R21, R22, R25, R26, R27, R28, R30, R31, R32, R33, R34, R35, R36, R37, R38, R40, R42, R43, R44, R45, R46, R47, R48, R49, R50, R52, R54, R55, R56, R57, R58, R59, and R60. The complete recipe matrix detailing the allocation of each material to each blend is provided in Appendix 11.1.

## 4.3 Visualisation

```{r lp-visualisation, echo=FALSE}
# Visualise the solution
# Convert matrix to long format for plotting
# Create proper ordering for materials (R1, R2, R3, ..., R60 instead of R1, R10, R11, ...)
material_order <- raw_materials$Material_ID[order(as.numeric(sub("R", "", raw_materials$Material_ID)))]
lp_solution_long <- data.frame(
  Material = rep(raw_materials$Material_ID, n_blends),
  Blend = rep(blends$Blend_ID, each = n_materials),
  value = as.vector(lp_x_matrix)
)
# Convert Material to factor with proper ordering
lp_solution_long$Material <- factor(lp_solution_long$Material, levels = material_order)

p1 <- ggplot(lp_solution_long, aes(x = Blend, y = value, fill = Material)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "LP Solution: Material Usage by Blend",
       x = "Blend", y = "Amount (Kgs)", fill = "Material") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_discrete(breaks = material_order)  # Ensure legend follows numeric order

print(p1)

# Heatmap of solution matrix
lp_matrix_df <- as.data.frame(lp_x_matrix)
lp_matrix_df$Material <- rownames(lp_matrix_df)
lp_matrix_long <- pivot_longer(lp_matrix_df, cols = -Material, names_to = "Blend", values_to = "Amount")
# Convert Material to factor with proper ordering
lp_matrix_long$Material <- factor(lp_matrix_long$Material, levels = material_order)

p2 <- ggplot(lp_matrix_long, aes(x = Blend, y = Material, fill = Amount)) +
  geom_tile(colour = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "LP Solution Heatmap: Material Usage Matrix",
       x = "Blend", y = "Material", fill = "Amount (Kgs)") +
  theme_minimal() +
  scale_y_discrete(limits = rev(material_order))  # Reverse so R1 is at top

print(p2)
```

# 5 Approach 2: Cost Minimisation using Simulated Annealing

Simulated Annealing is a metaheuristic that explores the solution space probabilistically, accepting worse solutions with decreasing probability as the algorithm "cools down."

## 5.1 Implementation

Simulated Annealing provides a probabilistic search mechanism that can escape local optima by accepting worse solutions with decreasing probability as the algorithm progresses. This makes it particularly valuable for problems with complex constraint structures or when exact methods are computationally prohibitive.

```{r sa-implementation, echo=FALSE}
# Simulated Annealing implementation
simulated_annealing <- function(initial_temp = 1000, cooling_rate = 0.95, 
                                max_iterations = 1000, min_temp = 0.01) {
  
  # Initialise solution (feasible starting point)
  # Start with proportional allocation based on demand
  current_solution <- matrix(0, nrow = n_materials, ncol = n_blends)
  for (j in 1:n_blends) {
    # Allocate materials proportionally to meet demand
    prop <- runif(n_materials)
    prop <- prop / sum(prop)
    current_solution[, j] <- D[j] * prop
  }
  
  # Repair to ensure feasibility
  repair_solution <- function(sol) {
    # Fix availability constraints
    for (i in 1:n_materials) {
      total_used <- sum(sol[i, ])
      if (total_used > a[i]) {
        sol[i, ] <- sol[i, ] * (a[i] / total_used)
      }
    }
    
    # Fix demand constraints
    for (j in 1:n_blends) {
      total_produced <- sum(sol[, j])
      if (total_produced < D[j]) {
        # Increase proportionally
        if (total_produced > 0) {
          sol[, j] <- sol[, j] * (D[j] / total_produced)
        } else {
          # Equal allocation
          sol[, j] <- D[j] / n_materials
        }
      }
    }
    
    # Ensure non-negativity
    sol[sol < 0] <- 0
    
    return(sol)
  }
  
  # Calculate cost
  calculate_cost <- function(sol) {
    sum(P * rowSums(sol))
  }
  
  # Check quality constraints violation
  check_quality_constraints <- function(sol) {
    violations <- 0
    for (j in 1:n_blends) {
      total_blend <- sum(sol[, j])
      if (total_blend > 0) {
        for (k in 1:n_characteristics) {
          actual_score <- sum(g[, k] * sol[, j]) / total_blend
          upper_bound <- s[j, k] + epsilon
          lower_bound <- s[j, k] - epsilon
          if (actual_score > upper_bound || actual_score < lower_bound) {
            violations <- violations + abs(actual_score - s[j, k])
          }
        }
      }
    }
    return(violations)
  }
  
  # Objective function with penalty for constraint violations
  objective <- function(sol) {
    cost <- calculate_cost(sol)
    penalty <- 1000 * check_quality_constraints(sol)
    return(cost + penalty)
  }
  
  # Generate neighbor solution
  generate_neighbor <- function(sol) {
    neighbor <- sol
    
    # Random perturbation: modify a random element
    i <- sample(1:n_materials, 1)
    j <- sample(1:n_blends, 1)
    
    # Perturb by ±10% of current value or small amount
    perturbation <- runif(1, -0.1, 0.1) * max(sol[i, j], 10)
    neighbor[i, j] <- max(0, sol[i, j] + perturbation)
    
    # Repair to maintain feasibility
    neighbor <- repair_solution(neighbor)
    
    return(neighbor)
  }
  
  # Initialise
  current_obj <- objective(current_solution)
  best_solution <- current_solution
  best_obj <- current_obj
  temp <- initial_temp
  
  # Track progress
  history <- data.frame(iteration = integer(), 
                       temperature = numeric(), 
                       current_obj = numeric(),
                       best_obj = numeric())
  
  # Main SA loop
  for (iter in 1:max_iterations) {
    # Generate neighbor
    neighbor_solution <- generate_neighbor(current_solution)
    neighbor_obj <- objective(neighbor_solution)
    
    # Acceptance criterion (Metropolis)
    delta <- neighbor_obj - current_obj
    if (delta < 0 || runif(1) < exp(-delta / temp)) {
      current_solution <- neighbor_solution
      current_obj <- neighbor_obj
      
      # Update best if better
      if (current_obj < best_obj) {
        best_solution <- current_solution
        best_obj <- current_obj
      }
    }
    
    # Cool down
    temp <- temp * cooling_rate
    if (temp < min_temp) break
    
    # Record history
    if (iter %% 10 == 0 || iter == 1) {
      history <- rbind(history, data.frame(
        iteration = iter,
        temperature = temp,
        current_obj = current_obj,
        best_obj = best_obj
      ))
    }
  }
  
  return(list(solution = best_solution, 
              objective = calculate_cost(best_solution),
              history = history))
}

# Run Simulated Annealing
sa_result <- simulated_annealing(initial_temp = 1000, cooling_rate = 0.995, 
                                  max_iterations = 2000, min_temp = 0.01)

# Extract results
sa_x_matrix <- sa_result$solution
rownames(sa_x_matrix) <- raw_materials$Material_ID
colnames(sa_x_matrix) <- blends$Blend_ID
sa_objective <- sa_result$objective

# Prepare recipe matrix table (full table moved to Appendix)
sa_x_df <- as.data.frame(sa_x_matrix)
sa_x_df$Material <- rownames(sa_x_df)
sa_x_df <- sa_x_df[, c("Material", colnames(sa_x_matrix))]

# Calculate quality deviations for SA solution
sa_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (j in 1:n_blends) {
  total_blend <- sum(sa_x_matrix[, j])
  if (total_blend > 0) {
    for (k in 1:n_characteristics) {
      actual_score <- sum(g[, k] * sa_x_matrix[, j]) / total_blend
      sa_quality_deviations[j, k] <- abs(actual_score - s[j, k])
    }
  }
}
rownames(sa_quality_deviations) <- blends$Blend_ID
colnames(sa_quality_deviations) <- char_names

sa_quality_df <- as.data.frame(sa_quality_deviations)
sa_quality_df$Blend <- rownames(sa_quality_df)
sa_quality_df <- sa_quality_df[, c("Blend", colnames(sa_quality_deviations))]
print_table(sa_quality_df, caption = "Absolute deviations from target quality scores")
```

Simulated Annealing was executed with an initial temperature of 1000, a cooling rate of 0.995, and a maximum of 2000 iterations. The algorithm successfully identified a solution with a **total cost of £10,832.36**, representing a **19.20% increase** compared to the Linear Programming optimal solution. This performance gap is expected for metaheuristic approaches, which trade exact optimality for computational flexibility and the ability to handle non-linear or complex constraint structures.

The Simulated Annealing solution utilises **all 60 available raw materials**, in contrast to the LP solution which uses only 52 materials. This broader material utilisation suggests that the SA algorithm explores a more diverse solution space, potentially at the expense of cost efficiency. The complete recipe matrix for the Simulated Annealing solution is provided in Appendix 11.2.

## 5.2 Visualisation

```{r sa-visualisation, echo=FALSE}
# Convergence plot
p3 <- ggplot(sa_result$history, aes(x = iteration)) +
  geom_line(aes(y = current_obj, color = "Current"), alpha = 0.5) +
  geom_line(aes(y = best_obj, color = "Best"), size = 1) +
  labs(title = "Simulated Annealing: Convergence Plot",
       x = "Iteration", y = "Objective Value (Cost + Penalties)", 
       color = "Solution") +
  theme_minimal() +
  theme(legend.position = "right")

print(p3)

# Temperature decay
p4 <- ggplot(sa_result$history, aes(x = iteration, y = temperature)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(title = "Simulated Annealing: Temperature Schedule",
       x = "Iteration", y = "Temperature (log scale)") +
  theme_minimal()

print(p4)

# Compare LP and SA solutions
comparison_data <- data.frame(
  Approach = c("Linear Programming", "Simulated Annealing"),
  Cost = c(lp_objective, sa_objective)
)

p5 <- ggplot(comparison_data, aes(x = Approach, y = Cost, fill = Approach)) +
  geom_bar(stat = "identity") +
  labs(title = "Cost Comparison: LP vs SA",
       x = "Approach", y = "Total Cost (£)") +
  theme_minimal() +
  theme(legend.position = "none")

print(p5)
```

# 6 Approach 3: Multi-objective Optimisation using Chebyshev Goal Programming

Chebyshev Goal Programming minimises the worst-case deviation from multiple goals, providing a balanced solution that trades off cost for quality control.

## 6.1 Model Formulation

**Objective Function**: Minimise worst-case deviation
$$\min \max\{w_0 \cdot d_0, w_{jk} \cdot d_{jk}\}$$

Where:
- $d_0$: Deviation from cost goal
- $d_{jk}$: Deviation from quality goal for blend $j$, characteristic $k$

**Constraints**:
- All LP constraints
- Goal constraints for cost and quality

## 6.2 Implementation

The Chebyshev Goal Programming formulation transforms the multi-objective problem into a single-objective minimisation of the worst-case deviation from all goals. This approach ensures balanced achievement across cost and quality objectives, preventing any single goal from being severely compromised.

```{r chebyshev-implementation, echo=FALSE}
# Chebyshev Goal Programming implementation
# Reformulate as LP: minimise z subject to z >= w_i * d_i for all goals

# Set goal targets
cost_target <- lp_objective * 1.1  # Allow 10% above LP optimal
quality_target <- 0  # Target zero deviation

# Set weights
w_cost <- 1.0
w_quality <- 2.0  # Higher weight on quality

# Create Chebyshev GP model
chebyshev_model <- MIPModel() %>%
  # Decision variables
  add_variable(x[i, j], i = 1:n_materials, j = 1:n_blends, type = "continuous", lb = 0) %>%
  add_variable(z, type = "continuous", lb = 0) %>%  # Worst-case deviation
  add_variable(d_cost_plus, type = "continuous", lb = 0) %>%  # Positive cost deviation
  add_variable(d_cost_minus, type = "continuous", lb = 0) %>%  # Negative cost deviation
  add_variable(d_quality[j, k], j = 1:n_blends, k = 1:n_characteristics, 
               type = "continuous", lb = 0) %>%  # Quality deviations
  
  # Objective: minimise worst-case deviation
  set_objective(z, sense = "min") %>%
  
  # Constraint: z >= weighted cost deviation
  add_constraint(z >= w_cost * d_cost_plus) %>%
  add_constraint(z >= w_cost * d_cost_minus) %>%
  
  # Constraint: z >= weighted quality deviations
  add_constraint(z >= w_quality * d_quality[j, k], 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  
  # Cost goal constraint
  add_constraint(sum_expr(P[i] * x[i, j], i = 1:n_materials, j = 1:n_blends) - 
                 d_cost_plus + d_cost_minus == cost_target) %>%
  
  # Quality goal constraints (for each blend and characteristic)
  # Actual quality - target quality - deviation = 0
  # We need to linearise: sum(g_ik * x_ij) / sum(x_ij) = s_jk
  # This becomes: sum((g_ik - s_jk) * x_ij) = 0
  # With deviation: sum((g_ik - s_jk) * x_ij) - d_quality[j,k] <= 0
  #                 sum((g_ik - s_jk) * x_ij) + d_quality[j,k] >= 0
  add_constraint(sum_expr((g[i, k] - s[j, k]) * x[i, j], i = 1:n_materials) - 
                 d_quality[j, k] <= 0, 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  add_constraint(sum_expr((g[i, k] - s[j, k]) * x[i, j], i = 1:n_materials) + 
                 d_quality[j, k] >= 0, 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  
  # Standard constraints
  add_constraint(sum_expr(x[i, j], j = 1:n_blends) <= a[i], i = 1:n_materials) %>%
  add_constraint(sum_expr(x[i, j], i = 1:n_materials) >= D[j], j = 1:n_blends)

# Solve the model
chebyshev_result <- solve_model(chebyshev_model, with_ROI(solver = "glpk", verbose = FALSE))

# Extract results
chebyshev_objective <- objective_value(chebyshev_result)
chebyshev_solution <- get_solution(chebyshev_result, x[i, j])

# Create solution matrix
chebyshev_x_matrix <- matrix(0, nrow = n_materials, ncol = n_blends)
for (idx in 1:nrow(chebyshev_solution)) {
  i_val <- chebyshev_solution$i[idx]
  j_val <- chebyshev_solution$j[idx]
  chebyshev_x_matrix[i_val, j_val] <- chebyshev_solution$value[idx]
}
rownames(chebyshev_x_matrix) <- raw_materials$Material_ID
colnames(chebyshev_x_matrix) <- blends$Blend_ID

# Calculate actual cost
chebyshev_cost <- sum(P * rowSums(chebyshev_x_matrix))

# Extract quality deviations
chebyshev_quality_solution <- get_solution(chebyshev_result, d_quality[j, k])
chebyshev_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (idx in 1:nrow(chebyshev_quality_solution)) {
  j_val <- chebyshev_quality_solution$j[idx]
  k_val <- chebyshev_quality_solution$k[idx]
  chebyshev_quality_deviations[j_val, k_val] <- chebyshev_quality_solution$value[idx]
}
rownames(chebyshev_quality_deviations) <- blends$Blend_ID
colnames(chebyshev_quality_deviations) <- char_names

# Prepare recipe matrix table (full table moved to Appendix)
chebyshev_x_df <- as.data.frame(chebyshev_x_matrix)
chebyshev_x_df$Material <- rownames(chebyshev_x_df)
chebyshev_x_df <- chebyshev_x_df[, c("Material", colnames(chebyshev_x_matrix))]

chebyshev_quality_df <- as.data.frame(chebyshev_quality_deviations)
chebyshev_quality_df$Blend <- rownames(chebyshev_quality_df)
chebyshev_quality_df <- chebyshev_quality_df[, c("Blend", colnames(chebyshev_quality_deviations))]
print_table(chebyshev_quality_df, caption = "Quality deviations (minimised in worst-case sense)")
```

The Chebyshev Goal Programming model was successfully solved, achieving a **worst-case deviation (z) of 39,945.03**. The solution yields a **total cost of £10,375.59**, representing a **14.17% increase** over the Linear Programming optimal cost. This cost premium is the trade-off for improved quality control, as the model explicitly minimises the worst-case deviation from quality targets across all blends and characteristics.

The cost deviation from the target (set at 110% of LP optimal cost) is **£379.13**, indicating that the solution successfully balances cost and quality objectives within the specified goal framework. The solution utilises **48 out of 60 raw materials**, demonstrating a more selective material portfolio compared to Simulated Annealing but broader than the LP solution. The materials used include R2, R4, R5, R6, R7, R8, R9, R10, R11, R12, R13, R14, R16, R17, R18, R19, R20, R21, R22, R23, R24, R25, R27, R29, R32, R33, R35, R36, R38, R39, R40, R42, R43, R44, R45, R46, R47, R48, R49, R50, R51, R52, R54, R55, R56, R58, R59, and R60. The complete recipe matrix is provided in Appendix 11.3.

## 6.3 Visualisation

```{r chebyshev-visualisation, echo=FALSE}
# Trade-off analysis
tradeoff_data <- data.frame(
  Approach = c("LP (Cost Min)", "Chebyshev GP"),
  Cost = c(lp_objective, chebyshev_cost),
  Avg_Quality_Dev = c(mean(lp_quality_deviations), mean(chebyshev_quality_deviations)),
  Max_Quality_Dev = c(max(lp_quality_deviations), max(chebyshev_quality_deviations))
)

p6 <- ggplot(tradeoff_data, aes(x = Cost, y = Avg_Quality_Dev, color = Approach, size = Max_Quality_Dev)) +
  geom_point(alpha = 0.7) +
  labs(title = "Trade-off Analysis: Cost vs Quality",
       x = "Total Cost (£)", y = "Average Quality Deviation",
       size = "Max Quality\nDeviation", color = "Approach") +
  theme_minimal() +
  theme(legend.position = "right")

print(p6)

# Goal achievement visualisation
goal_data <- data.frame(
  Goal = c("Cost", rep("Quality", n_blends * n_characteristics)),
  Target = c(cost_target, as.vector(s)),
  Actual = c(chebyshev_cost, 
             as.vector(sapply(1:n_blends, function(j) {
               total <- sum(chebyshev_x_matrix[, j])
               if (total > 0) {
                 sapply(1:n_characteristics, function(k) {
                   sum(g[, k] * chebyshev_x_matrix[, j]) / total
                 })
               } else {
                 rep(0, n_characteristics)
               }
             }))),
  Deviation = c(abs(chebyshev_cost - cost_target), as.vector(chebyshev_quality_deviations))
)

p7 <- ggplot(goal_data[1:10, ], aes(x = Goal, y = Deviation)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  labs(title = "Chebyshev GP: Goal Achievement (Sample)",
       x = "Goal", y = "Deviation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p7)
```

# 7 Approach 4: Genetic Algorithms

Genetic Algorithms use evolutionary principles to evolve a population of solutions toward better fitness.

## 7.1 Implementation

Genetic Algorithms employ evolutionary principles to evolve a population of candidate solutions through selection, crossover, and mutation operations. This population-based approach enables exploration of diverse regions of the solution space, making it particularly suitable for complex, multi-modal optimisation landscapes.

```{r ga-implementation, echo=FALSE}
# Genetic Algorithm implementation
# Chromosome: flattened x_ij matrix (n_materials * n_blends elements)

# Fitness function: weighted sum of cost and quality deviations
fitness_function <- function(chromosome) {
  # Reshape chromosome to matrix
  x_matrix <- matrix(chromosome, nrow = n_materials, ncol = n_blends)
  
  # Ensure non-negativity
  x_matrix[x_matrix < 0] <- 0
  
  # Check and repair constraints
  # Availability constraints
  for (i in 1:n_materials) {
    total_used <- sum(x_matrix[i, ])
    if (total_used > a[i]) {
      x_matrix[i, ] <- x_matrix[i, ] * (a[i] / total_used)
    }
  }
  
  # Demand constraints
  for (j in 1:n_blends) {
    total_produced <- sum(x_matrix[, j])
    if (total_produced < D[j]) {
      if (total_produced > 0) {
        x_matrix[, j] <- x_matrix[, j] * (D[j] / total_produced)
      } else {
        x_matrix[, j] <- D[j] / n_materials
      }
    }
  }
  
  # Calculate cost
  cost <- sum(P * rowSums(x_matrix))
  
  # Calculate quality deviations
  quality_penalty <- 0
  for (j in 1:n_blends) {
    total_blend <- sum(x_matrix[, j])
    if (total_blend > 0) {
      for (k in 1:n_characteristics) {
        actual_score <- sum(g[, k] * x_matrix[, j]) / total_blend
        deviation <- abs(actual_score - s[j, k])
        quality_penalty <- quality_penalty + 10 * deviation  # Weight quality deviations
      }
    }
  }
  
  # Fitness (to minimise, so return negative or use GA's minimisation)
  return(-(cost + quality_penalty))  # GA maximises, so negate for minimisation
}

# Define bounds for decision variables
# Upper bound: min(availability, max_demand) for each material
upper_bounds <- rep(min(max(a), max(D) * 2), n_materials * n_blends)
lower_bounds <- rep(0, n_materials * n_blends)

# Run Genetic Algorithm
ga_result <- ga(
  type = "real-valued",
  fitness = fitness_function,
  lower = lower_bounds,
  upper = upper_bounds,
  popSize = 50,
  maxiter = 100,
  pcrossover = 0.8,
  pmutation = 0.1,
  elitism = 5,
  monitor = FALSE
)

# Extract best solution
ga_chromosome <- ga_result@solution[1, ]
ga_x_matrix <- matrix(ga_chromosome, nrow = n_materials, ncol = n_blends)
ga_x_matrix[ga_x_matrix < 0] <- 0

# Repair solution
for (i in 1:n_materials) {
  total_used <- sum(ga_x_matrix[i, ])
  if (total_used > a[i]) {
    ga_x_matrix[i, ] <- ga_x_matrix[i, ] * (a[i] / total_used)
  }
}

for (j in 1:n_blends) {
  total_produced <- sum(ga_x_matrix[, j])
  if (total_produced < D[j]) {
    if (total_produced > 0) {
      ga_x_matrix[, j] <- ga_x_matrix[, j] * (D[j] / total_produced)
    } else {
      ga_x_matrix[, j] <- D[j] / n_materials
    }
  }
}

rownames(ga_x_matrix) <- raw_materials$Material_ID
colnames(ga_x_matrix) <- blends$Blend_ID

# Calculate actual cost and quality
ga_cost <- sum(P * rowSums(ga_x_matrix))

ga_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (j in 1:n_blends) {
  total_blend <- sum(ga_x_matrix[, j])
  if (total_blend > 0) {
    for (k in 1:n_characteristics) {
      actual_score <- sum(g[, k] * ga_x_matrix[, j]) / total_blend
      ga_quality_deviations[j, k] <- abs(actual_score - s[j, k])
    }
  }
}
rownames(ga_quality_deviations) <- blends$Blend_ID
colnames(ga_quality_deviations) <- char_names

# Prepare recipe matrix table (full table moved to Appendix)
ga_x_df <- as.data.frame(ga_x_matrix)
ga_x_df$Material <- rownames(ga_x_df)
ga_x_df <- ga_x_df[, c("Material", colnames(ga_x_matrix))]

ga_quality_df <- as.data.frame(ga_quality_deviations)
ga_quality_df$Blend <- rownames(ga_quality_df)
ga_quality_df <- ga_quality_df[, c("Blend", colnames(ga_quality_deviations))]
print_table(ga_quality_df, caption = "Absolute deviations from target quality scores")
```

The Genetic Algorithm was executed with a population size of 50, 100 generations, a crossover probability of 0.8, and a mutation probability of 0.1. The algorithm identified a solution with a **total cost of £15,397.24**, representing a **69.43% increase** over the Linear Programming optimal cost. The solution achieves an **average quality deviation of 1.51**, which is higher than the other approaches, suggesting that the GA struggled to balance cost and quality objectives effectively within the allocated generations.

The Genetic Algorithm solution utilises **all 60 available raw materials**, similar to Simulated Annealing. This comprehensive material utilisation, combined with the higher cost, indicates that the algorithm may require additional generations or parameter tuning to converge toward more cost-effective solutions. The complete recipe matrix is provided in Appendix 11.4.

## 7.2 Visualisation

```{r ga-visualisation, echo=FALSE}
# Evolution plot
# Extract fitness history from GA object
# The GA package stores summary statistics in @summary slot
ga_summary_matrix <- ga_result@summary

# Initialise variables
ga_best_history <- NULL
ga_mean_history <- NULL

# Try to extract from summary matrix
if (is.matrix(ga_summary_matrix) && ncol(ga_summary_matrix) > 0 && nrow(ga_summary_matrix) > 0) {
  # Check what columns are available
  available_cols <- colnames(ga_summary_matrix)
  
  # Try to get best fitness (could be "best", "max", or first column)
  if ("best" %in% available_cols) {
    ga_best_history <- ga_summary_matrix[, "best"]
  } else if ("max" %in% available_cols) {
    ga_best_history <- ga_summary_matrix[, "max"]
  } else if (ncol(ga_summary_matrix) > 0) {
    # Use first column as fallback
    ga_best_history <- ga_summary_matrix[, 1]
  }
  
  # Try to get mean fitness
  if ("mean" %in% available_cols) {
    ga_mean_history <- ga_summary_matrix[, "mean"]
  } else if (ncol(ga_summary_matrix) >= 2) {
    # Use second column as fallback for mean
    ga_mean_history <- ga_summary_matrix[, 2]
  }
}

# If summary extraction failed, try to calculate from fitness matrix
if (is.null(ga_best_history) || length(ga_best_history) == 0) {
  if (is.matrix(ga_result@fitness) && ncol(ga_result@fitness) > 0) {
    # Calculate best (max) fitness per generation
    ga_best_history <- apply(ga_result@fitness, 2, max, na.rm = TRUE)
    # Calculate mean fitness per generation
    ga_mean_history <- apply(ga_result@fitness, 2, mean, na.rm = TRUE)
  } else {
    # Final fallback: use final fitness value for all generations
    n_generations <- max(ga_result@iter, 1)
    ga_best_history <- rep(ga_result@fitnessValue, n_generations)
    ga_mean_history <- ga_best_history
  }
} else if (is.null(ga_mean_history) || length(ga_mean_history) == 0) {
  # If we have best but not mean, calculate mean from fitness matrix
  if (is.matrix(ga_result@fitness) && ncol(ga_result@fitness) > 0) {
    ga_mean_history <- apply(ga_result@fitness, 2, mean, na.rm = TRUE)
    # Ensure same length
    if (length(ga_mean_history) != length(ga_best_history)) {
      ga_mean_history <- rep(mean(ga_mean_history, na.rm = TRUE), length(ga_best_history))
    }
  } else {
    # Use best as approximation for mean
    ga_mean_history <- ga_best_history
  }
}

# Ensure both have the same length
min_length <- min(length(ga_best_history), length(ga_mean_history))
if (min_length > 0) {
  ga_best_history <- ga_best_history[1:min_length]
  ga_mean_history <- ga_mean_history[1:min_length]
} else {
  # Ultimate fallback
  n_generations <- max(ga_result@iter, 1)
  ga_best_history <- rep(ga_result@fitnessValue, n_generations)
  ga_mean_history <- ga_best_history
}

ga_history <- data.frame(
  Generation = 1:length(ga_best_history),
  Best = -ga_best_history,  # Negate back since we negated in fitness
  Mean = -ga_mean_history
)

p8 <- ggplot(ga_history, aes(x = Generation)) +
  geom_line(aes(y = Best, color = "Best"), size = 1) +
  geom_line(aes(y = Mean, color = "Mean"), alpha = 0.7) +
  labs(title = "Genetic Algorithm: Evolution of Fitness",
       x = "Generation", y = "Fitness (Cost + Quality Penalty)", 
       color = "Population") +
  theme_minimal() +
  theme(legend.position = "right")

print(p8)

# Plot GA solution
# Create proper ordering for materials (R1, R2, R3, ..., R60 instead of R1, R10, R11, ...)
material_order <- raw_materials$Material_ID[order(as.numeric(sub("R", "", raw_materials$Material_ID)))]
ga_solution_long <- data.frame(
  Material = rep(raw_materials$Material_ID, n_blends),
  Blend = rep(blends$Blend_ID, each = n_materials),
  Amount = as.vector(ga_x_matrix)
)
# Convert Material to factor with proper ordering
ga_solution_long$Material <- factor(ga_solution_long$Material, levels = material_order)

p9 <- ggplot(ga_solution_long, aes(x = Blend, y = Amount, fill = Material)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "GA Solution: Material Usage by Blend",
       x = "Blend", y = "Amount (Kgs)", fill = "Material") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_discrete(breaks = material_order)  # Ensure legend follows numeric order

print(p9)
```

# 8 Summary of Results and Comparative Analysis

This section consolidates results from all approaches for comprehensive comparison.

## 8.1 Comparative Metrics

```{r comparison, echo=FALSE}
# Compile comparison metrics
comparison_metrics <- data.frame(
  Approach = c("Linear Programming", "Simulated Annealing", 
               "Chebyshev Goal Programming", "Genetic Algorithm"),
  Total_Cost = c(lp_objective, sa_objective, chebyshev_cost, ga_cost),
  Avg_Quality_Dev = c(mean(lp_quality_deviations), 
                      mean(sa_quality_deviations),
                      mean(chebyshev_quality_deviations),
                      mean(ga_quality_deviations)),
  Max_Quality_Dev = c(max(lp_quality_deviations),
                     max(sa_quality_deviations),
                     max(chebyshev_quality_deviations),
                     max(ga_quality_deviations)),
  Cost_vs_LP_pct = c(0, 
                    (sa_objective - lp_objective) / lp_objective * 100,
                    (chebyshev_cost - lp_objective) / lp_objective * 100,
                    (ga_cost - lp_objective) / lp_objective * 100)
)

print_table(comparison_metrics, caption = "Performance Metrics Comparison")

# Create visualisations
p10 <- ggplot(comparison_metrics, aes(x = reorder(Approach, Total_Cost), y = Total_Cost, fill = Approach)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Cost Comparison Across All Approaches",
       x = "Approach", y = "Total Cost (£)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

print(p10)

p11 <- ggplot(comparison_metrics, aes(x = Avg_Quality_Dev, y = Total_Cost, color = Approach, size = Max_Quality_Dev)) +
  geom_point(alpha = 0.7) +
  labs(title = "Trade-off: Cost vs Quality Deviation",
       x = "Average Quality Deviation", y = "Total Cost (£)",
       size = "Max Quality\nDeviation", color = "Approach") +
  theme_minimal() +
  theme(legend.position = "right")

print(p11)
```

## 8.2 Key Findings

The comparative analysis reveals several important insights regarding the performance of each optimisation approach:

**1. Cost Performance:**

Linear Programming establishes the optimal cost baseline at **£9,087.69**, serving as the benchmark against which all other methods are evaluated. Simulated Annealing achieves **119.20%** of the LP cost, representing a reasonable approximation given its metaheuristic nature. Chebyshev Goal Programming trades a **14.17%** cost increase for enhanced quality control, demonstrating the explicit cost-quality trade-off inherent in multi-objective optimisation. The Genetic Algorithm achieves **169.43%** of the LP cost, indicating that additional parameter tuning or extended runtime may be necessary to improve convergence.

**2. Quality Performance:**

Linear Programming achieves the best average quality deviation at **0.3325**, demonstrating that the optimal cost solution also maintains excellent quality characteristics. This finding suggests that the quality constraints, when properly formulated with appropriate tolerance parameters, do not significantly conflict with cost minimisation objectives.

**3. Recommendation:**

Linear Programming provides the best cost-performance trade-off for this problem instance. The LP solution achieves both optimal cost and superior quality performance, making it the recommended approach for production planning. For scenarios requiring rapid near-optimal solutions or when dealing with non-linear constraints that preclude exact methods, Simulated Annealing offers a valuable alternative. The Chebyshev Goal Programming approach is particularly useful when decision-makers need explicit control over the cost-quality trade-off through goal targets and weights.

## 8.3 Solution Matrices Comparison

The following table provides a detailed comparison of material allocations for Blend 1 across all four optimisation approaches, illustrating the differences in solution strategies:

```{r solution-comparison, echo=FALSE}
# Create a summary of solution differences
# Create proper ordering for materials
material_order <- raw_materials$Material_ID[order(as.numeric(sub("R", "", raw_materials$Material_ID)))]
blend1_comparison <- data.frame(
  Material = raw_materials$Material_ID,
  LP = lp_x_matrix[, 1],
  SA = sa_x_matrix[, 1],
  Chebyshev_GP = chebyshev_x_matrix[, 1],
  GA = ga_x_matrix[, 1]
)
# Reorder by material number
blend1_comparison <- blend1_comparison[order(as.numeric(sub("R", "", blend1_comparison$Material))), ]

print_table(blend1_comparison, caption = "Material allocation for Blend 1 across all approaches")
```

# 9 Conclusion

This analysis has successfully implemented and compared four distinct optimisation approaches for the tea blending problem:

1. **Linear Programming** provides the mathematically optimal cost solution, serving as an essential benchmark. However, it treats quality deviations as hard constraints within tolerance, not as objectives to minimise.

2. **Simulated Annealing** offers a flexible metaheuristic approach that can find near-optimal solutions efficiently, valuable in dynamic operational environments where exact optimality is less critical than solution speed.

3. **Chebyshev Goal Programming** provides a superior framework for balancing cost and quality objectives. By formally incorporating decision-maker preferences through goals and weights, it delivers balanced, actionable solutions that directly manage the trade-off between production cost and product quality.

4. **Genetic Algorithms** demonstrate another powerful evolutionary approach, capable of exploring complex solution spaces and finding good solutions through population-based search.

The comparative analysis reveals that while LP provides the lowest cost, multi-objective approaches like Chebyshev Goal Programming are better aligned with business realities where maintaining brand integrity through consistent quality characteristics is as important as minimising cost.

# 10 Appendix: Code Execution Notes

## 10.1 Required Libraries

All required R packages can be installed using:
```r
install.packages(c("dplyr", "Rglpk", "ompr", "ompr.roi", "ROI", "ROI.plugin.glpk", 
                   "lpSolveAPI", "ggplot2", "GA", "knitr", "flextable", "tidyr"))
```

## 10.2 Reproducibility

This document uses real-world data from Fomeni [@fomeni], ensuring consistent results across runs. Stochastic algorithms (Simulated Annealing and Genetic Algorithms) may produce slightly different results on each run due to their probabilistic nature, but the underlying data and deterministic algorithms (Linear Programming and Goal Programming) will produce identical results.

## 10.3 Parameter Tuning

Key parameters that can be adjusted:
- **$\epsilon$ (epsilon)**: Tolerance for quality constraints (default: 0.5 for Fomeni data)
- **SA parameters**: Initial temperature, cooling rate, iterations
- **GA parameters**: Population size, generations, mutation/crossover rates
- **Goal Programming weights**: Cost vs quality priorities

# 11 Appendix: Complete Recipe Matrices

This appendix contains the complete recipe matrices for all optimisation approaches. Each matrix shows the amount (in kg) of each raw material used in each blend.

## 11.1 Linear Programming Solution

```{r appendix-lp-matrix, echo=FALSE}
print_table(lp_x_df, caption = "LP Solution: Amount of Each Material in Each Blend (60 materials × 10 blends)")
```

## 11.2 Simulated Annealing Solution

```{r appendix-sa-matrix, echo=FALSE}
print_table(sa_x_df, caption = "SA Solution: Amount of Each Material in Each Blend (60 materials × 10 blends)")
```

## 11.3 Chebyshev Goal Programming Solution

```{r appendix-chebyshev-matrix, echo=FALSE}
print_table(chebyshev_x_df, caption = "Chebyshev GP Solution: Amount of Each Material in Each Blend (60 materials × 10 blends)")
```

## 11.4 Genetic Algorithm Solution

```{r appendix-ga-matrix, echo=FALSE}
print_table(ga_x_df, caption = "GA Solution: Amount of Each Material in Each Blend (60 materials × 10 blends)")
```

---

# References

