---
title: "Tea Blending Optimisation: A Comprehensive Analysis"
subtitle: "STA5074Z Prescriptive Analytics"
author: "Tinotenda Muzambi - MZMTIN002"
date: "`r Sys.Date()`"
bibliography: references.bib
csl: apa-numeric-superscript-brackets.csl
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
  pdf_document:
    toc: true
    latex_engine: xelatex  # Support Unicode characters
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6, fig.align = "center")

# Load required libraries
library(dplyr)
library(Rglpk)
library(ompr)
library(ompr.roi)
library(ROI)
library(ROI.plugin.glpk)
library(lpSolveAPI)
library(ggplot2)
library(GA)
library(knitr)
library(flextable) 
library(tidyr)

# Helper function to print tables 
print_table <- function(data, caption = "") {
  ft <- flextable(data) %>%
    set_caption(caption) %>%
    theme_zebra(odd_header = "#CFCFCF", odd_body = "#EFEFEF") %>%
    autofit()
  
  if (knitr::is_latex_output()) {
    ft <- ft %>%
      fontsize(size = 9) %>%
      set_table_properties(layout = "autofit")
  }
  
  return(ft)
}
```

# 1 Introduction
Blending problems are a classic challenge in operations research, arising in diverse fields such as food processing, metallurgy, and energy. In the tea industry, blending is crucial to achieve consistent product quality, balance raw material costs, and adapt to fluctuating supplies. Every cup of tea on the supermarket shelf is the result of a carefully orchestrated process: different grades and types of tea leaves from various origins are combined to meet precise targets for taste, color, aroma, and other quality parameters, all while minimising production costs and ensuring supply constraints are not violated.

In the context of tea manufacturing, the blending optimisation problem involves selecting the optimal composition of raw teas (inputs) to satisfy desired blend qualities for multiple final blends. This not only impacts the sensory profile and brand identity of each product, but also has substantial operational and economic implications. By applying advanced computational techniques, such as mathematical programming and metaheuristics, producers can automate and enhance the blending decision-making process, yielding better margins, consistent quality, and greater responsiveness to market demands.

Below, we formally define the tea blending optimisation problem and systematically explore several approaches for solving it. The problem requires determining the least-cost combination of raw materials to produce final blends that meet specific quality characteristics, as formalised by Fomeni [@fomeni].

The primary objectives are:

1. To implement and solve a cost-minimisation model using Linear Programming (LP).
2. To solve the same problem using Simulated Annealing (SA) metaheuristic and compare performance.
3. To formulate and solve a multi-objective model using Chebyshev Goal Programming.
4. To explore Genetic Algorithms (GAs) as an alternative solution approach.

# 2 Problem Formulation and Data Preparation

## 2.1 Notations and Parameters

- **Indices**:
  - $i$: Index for raw materials
  - $j$: Index for final blends
  - $k$: Index for quality characteristics

- **Parameters**:
  - $P_i$: Unit cost of raw material $i$ (£/Kg)
  - $D_j$: Weekly demand for final blend $j$ (Kgs)
  - $a_i$: Weekly availability of raw material $i$ (Kgs)
  - $g_{ik}$: Grade of characteristic $k$ in one unit of raw material $i$
  - $s_{jk}$: Required target score of characteristic $k$ in one unit of final blend $j$

- **Decision Variables**:
  - $x_{ij}$: Amount (in Kgs) of raw material $i$ to be used in final blend $j$

## 2.2 Sample Data Generation

Based on the structure described in Fomeni [@fomeni], we generate sample data for raw materials and blends.

```{r data-generation}
# Set seed for reproducibility
set.seed(2025)

# Define number of raw materials, blends, and quality characteristics
n_materials <- 6
n_blends <- 4
n_characteristics <- 3

# Quality characteristic names
char_names <- c("Brightness", "Colour", "Thickness")

# Generate raw materials data
raw_materials <- data.frame(
  Material_ID = paste0("RM", 1:n_materials),
  Cost = round(runif(n_materials, 2.5, 8.0), 2),
  Availability = round(runif(n_materials, 150, 400), 0),
  stringsAsFactors = FALSE
)

# Generate quality characteristics for raw materials (g_ik)
quality_matrix <- matrix(
  round(runif(n_materials * n_characteristics, 1, 10), 2),
  nrow = n_materials,
  ncol = n_characteristics
)
colnames(quality_matrix) <- char_names
raw_materials <- cbind(raw_materials, quality_matrix)

# Generate blends
blends <- data.frame(
  Blend_ID = paste0("Blend", 1:n_blends),
  Demand = round(runif(n_blends, 50, 200), 0),
  stringsAsFactors = FALSE
)

# Generate target quality scores for blends (s_jk)
target_matrix <- matrix(
  round(runif(n_blends * n_characteristics, 1, 10), 2),
  nrow = n_blends,
  ncol = n_characteristics
)
colnames(target_matrix) <- char_names
blends <- cbind(blends, target_matrix)

# Set tolerance parameter (epsilon)
epsilon <- 1.0

# Display the data
cat("=== Raw Materials Data (Table 2) ===\n")
print_table(raw_materials, caption = "Raw Materials: Costs, Availability, and Quality Characteristics")

cat("\n=== Blends Data (Table 3) ===\n")
print_table(blends, caption = "Blends: Demand and Target Quality Scores")

# Store data in convenient format for later use
# For ompr, we need numeric-indexed vectors (not named vectors)
P <- as.numeric(raw_materials$Cost)  # Cost vector indexed by i = 1:n_materials
a <- as.numeric(raw_materials$Availability)  # Availability vector indexed by i = 1:n_materials
D <- as.numeric(blends$Demand)  # Demand vector indexed by j = 1:n_blends

# Quality matrices: g[i,k] and s[j,k] indexed numerically
g <- as.matrix(raw_materials[, char_names])  # g[i,k] where i = 1:n_materials, k = 1:n_characteristics
s <- as.matrix(blends[, char_names])  # s[j,k] where j = 1:n_blends, k = 1:n_characteristics

# Keep named versions for display purposes
P_named <- raw_materials$Cost
names(P_named) <- raw_materials$Material_ID
a_named <- raw_materials$Availability
names(a_named) <- raw_materials$Material_ID
D_named <- blends$Demand
names(D_named) <- blends$Blend_ID
g_named <- as.matrix(raw_materials[, char_names])
rownames(g_named) <- raw_materials$Material_ID
s_named <- as.matrix(blends[, char_names])
rownames(s_named) <- blends$Blend_ID
```

# 3 Approach 1: Cost Minimisation using Linear Programming

This section implements the cost-minimisation model using Linear Programming with parametric relaxation of the characteristic-matching constraints. This involves implementing a cost-minimisation LP model that finds the optimal combination of raw materials to produce blends meeting demand and quality constraints. This implementation uses Rglpk which provides a straightforward matrix-based formulation that is well-suited for this problem type.

## 3.1 Model Formulation

**Objective Function**: Minimise total cost
$$\min \sum_{i} \sum_{j} P_i x_{ij}$$

**Subject to Constraints**:

1. Raw Material Availability: $\sum_{j} x_{ij} \le a_i \quad \forall i$
2. Blend Demand: $\sum_{i} x_{ij} \ge D_j \quad \forall j$
3. Characteristic Score Relaxation:
   - $\sum_{i} (g_{ik} - (s_{jk} + \epsilon_{jk})) x_{ij} \le 0 \quad \forall j, k$
   - $\sum_{i} (g_{ik} - (s_{jk} - \epsilon_{jk})) x_{ij} \ge 0 \quad \forall j, k$
4. Non-negativity: $x_{ij} \ge 0 \quad \forall i, j$

## 3.2 Implementation

```{r lp-implementation}
# Implementation using Rglpk

# Number of decision variables: n_materials * n_blends (x[i,j] for all i,j)
n_vars <- n_materials * n_blends

# Objective function: minimize cost
# Objective coefficients: P[i] for each x[i,j]
# Variable ordering: x[1,1], x[1,2], ..., x[1,n_blends], x[2,1], ..., x[n_materials,n_blends]
obj_coef <- rep(P, n_blends)  # Repeat P for each blend

# Build constraint matrix
# Constraints in order:
# 1. Availability: sum_j x[i,j] <= a[i] for each i (n_materials constraints)
# 2. Demand: sum_i x[i,j] >= D[j] for each j (n_blends constraints)  
# 3. Quality upper: sum_i (g[i,k] - (s[j,k] + eps)) * x[i,j] <= 0 (n_blends * n_characteristics)
# 4. Quality lower: sum_i (g[i,k] - (s[j,k] - eps)) * x[i,j] >= 0 (n_blends * n_characteristics)

n_avail_constraints <- n_materials
n_demand_constraints <- n_blends
n_quality_constraints <- 2 * n_blends * n_characteristics
n_constraints <- n_avail_constraints + n_demand_constraints + n_quality_constraints

# Initialize constraint matrix
constraint_matrix <- matrix(0, nrow = n_constraints, ncol = n_vars)

# RHS and direction vectors
rhs <- numeric(n_constraints)
direction <- character(n_constraints)

# Variable indexing function: x[i,j] is at column (i-1)*n_blends + j
get_col <- function(i, j) {
  (i - 1) * n_blends + j
}

# 1. Availability constraints: sum_j x[i,j] <= a[i]
row <- 0
for (i in 1:n_materials) {
  row <- row + 1
  for (j in 1:n_blends) {
    col <- get_col(i, j)
    constraint_matrix[row, col] <- 1
  }
  rhs[row] <- a[i]
  direction[row] <- "<="
}

# 2. Demand constraints: sum_i x[i,j] >= D[j]
for (j in 1:n_blends) {
  row <- row + 1
  for (i in 1:n_materials) {
    col <- get_col(i, j)
    constraint_matrix[row, col] <- 1
  }
  rhs[row] <- D[j]
  direction[row] <- ">="
}

# 3. Quality constraints (upper bound): sum_i (g[i,k] - (s[j,k] + eps)) * x[i,j] <= 0
for (j in 1:n_blends) {
  for (k in 1:n_characteristics) {
    row <- row + 1
    for (i in 1:n_materials) {
      col <- get_col(i, j)
      constraint_matrix[row, col] <- g[i, k] - (s[j, k] + epsilon)
    }
    rhs[row] <- 0
    direction[row] <- "<="
  }
}

# 4. Quality constraints (lower bound): sum_i (g[i,k] - (s[j,k] - eps)) * x[i,j] >= 0
for (j in 1:n_blends) {
  for (k in 1:n_characteristics) {
    row <- row + 1
    for (i in 1:n_materials) {
      col <- get_col(i, j)
      constraint_matrix[row, col] <- g[i, k] - (s[j, k] - epsilon)
    }
    rhs[row] <- 0
    direction[row] <- ">="
  }
}

# Variable bounds (non-negativity)
bounds <- list(lower = list(ind = 1:n_vars, val = rep(0, n_vars)),
                upper = list(ind = 1:n_vars, val = rep(Inf, n_vars)))

# Solve the LP
result <- Rglpk_solve_LP(obj = obj_coef, 
                         mat = constraint_matrix, 
                         dir = direction, 
                         rhs = rhs,
                         bounds = bounds,
                         max = FALSE)

# Extract results
lp_objective <- result$optimum
lp_solution_vector <- result$solution

# Reshape solution vector into matrix
# Solution vector is: x[1,1], x[1,2], ..., x[1,n_blends], x[2,1], ..., x[n_materials,n_blends]
lp_x_matrix <- matrix(lp_solution_vector, nrow = n_materials, ncol = n_blends, byrow = TRUE)
rownames(lp_x_matrix) <- raw_materials$Material_ID
colnames(lp_x_matrix) <- blends$Blend_ID

cat(sprintf("\n=== LP Solution Results ===\n"))
cat(sprintf("Optimal Total Cost: £%.2f\n", lp_objective))

cat("\n=== Optimal Recipe Matrix (x_ij in Kgs) ===\n")
print_table(as.data.frame(lp_x_matrix), caption = "LP Solution: Amount of Each Material in Each Blend")

# Calculate quality deviations for LP solution
lp_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (j in 1:n_blends) {
  total_blend <- sum(lp_x_matrix[, j])
  if (total_blend > 0) {
    for (k in 1:n_characteristics) {
      actual_score <- sum(g[, k] * lp_x_matrix[, j]) / total_blend
      lp_quality_deviations[j, k] <- abs(actual_score - s[j, k])
    }
  }
}
rownames(lp_quality_deviations) <- blends$Blend_ID
colnames(lp_quality_deviations) <- char_names

cat("\n=== Quality Deviations from Targets ===\n")
print_table(as.data.frame(lp_quality_deviations), caption = "Absolute deviations from target quality scores")
```

## 3.3 Visualisation

```{r lp-visualisation}
# Visualise the solution
# Convert matrix to long format for plotting
lp_solution_long <- data.frame(
  Material = rep(raw_materials$Material_ID, n_blends),
  Blend = rep(blends$Blend_ID, each = n_materials),
  value = as.vector(lp_x_matrix)
)

p1 <- ggplot(lp_solution_long, aes(x = Blend, y = value, fill = Material)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "LP Solution: Material Usage by Blend",
       x = "Blend", y = "Amount (Kgs)", fill = "Material") +
  theme_minimal() +
  theme(legend.position = "right")

print(p1)

# Heatmap of solution matrix
lp_matrix_df <- as.data.frame(lp_x_matrix)
lp_matrix_df$Material <- rownames(lp_matrix_df)
lp_matrix_long <- pivot_longer(lp_matrix_df, cols = -Material, names_to = "Blend", values_to = "Amount")

p2 <- ggplot(lp_matrix_long, aes(x = Blend, y = Material, fill = Amount)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "LP Solution Heatmap: Material Usage Matrix",
       x = "Blend", y = "Material", fill = "Amount (Kgs)") +
  theme_minimal()

print(p2)
```

# 4 Approach 2: Cost Minimisation using Simulated Annealing

Simulated Annealing is a metaheuristic that explores the solution space probabilistically, accepting worse solutions with decreasing probability as the algorithm "cools down."

## 4.1 Implementation

```{r sa-implementation}
# Simulated Annealing implementation
simulated_annealing <- function(initial_temp = 1000, cooling_rate = 0.95, 
                                max_iterations = 1000, min_temp = 0.01) {
  
  # Initialise solution (feasible starting point)
  # Start with proportional allocation based on demand
  current_solution <- matrix(0, nrow = n_materials, ncol = n_blends)
  for (j in 1:n_blends) {
    # Allocate materials proportionally to meet demand
    prop <- runif(n_materials)
    prop <- prop / sum(prop)
    current_solution[, j] <- D[j] * prop
  }
  
  # Repair to ensure feasibility
  repair_solution <- function(sol) {
    # Fix availability constraints
    for (i in 1:n_materials) {
      total_used <- sum(sol[i, ])
      if (total_used > a[i]) {
        sol[i, ] <- sol[i, ] * (a[i] / total_used)
      }
    }
    
    # Fix demand constraints
    for (j in 1:n_blends) {
      total_produced <- sum(sol[, j])
      if (total_produced < D[j]) {
        # Increase proportionally
        if (total_produced > 0) {
          sol[, j] <- sol[, j] * (D[j] / total_produced)
        } else {
          # Equal allocation
          sol[, j] <- D[j] / n_materials
        }
      }
    }
    
    # Ensure non-negativity
    sol[sol < 0] <- 0
    
    return(sol)
  }
  
  # Calculate cost
  calculate_cost <- function(sol) {
    sum(P * rowSums(sol))
  }
  
  # Check quality constraints violation
  check_quality_constraints <- function(sol) {
    violations <- 0
    for (j in 1:n_blends) {
      total_blend <- sum(sol[, j])
      if (total_blend > 0) {
        for (k in 1:n_characteristics) {
          actual_score <- sum(g[, k] * sol[, j]) / total_blend
          upper_bound <- s[j, k] + epsilon
          lower_bound <- s[j, k] - epsilon
          if (actual_score > upper_bound || actual_score < lower_bound) {
            violations <- violations + abs(actual_score - s[j, k])
          }
        }
      }
    }
    return(violations)
  }
  
  # Objective function with penalty for constraint violations
  objective <- function(sol) {
    cost <- calculate_cost(sol)
    penalty <- 1000 * check_quality_constraints(sol)
    return(cost + penalty)
  }
  
  # Generate neighbor solution
  generate_neighbor <- function(sol) {
    neighbor <- sol
    
    # Random perturbation: modify a random element
    i <- sample(1:n_materials, 1)
    j <- sample(1:n_blends, 1)
    
    # Perturb by ±10% of current value or small amount
    perturbation <- runif(1, -0.1, 0.1) * max(sol[i, j], 10)
    neighbor[i, j] <- max(0, sol[i, j] + perturbation)
    
    # Repair to maintain feasibility
    neighbor <- repair_solution(neighbor)
    
    return(neighbor)
  }
  
  # Initialise
  current_obj <- objective(current_solution)
  best_solution <- current_solution
  best_obj <- current_obj
  temp <- initial_temp
  
  # Track progress
  history <- data.frame(iteration = integer(), 
                       temperature = numeric(), 
                       current_obj = numeric(),
                       best_obj = numeric())
  
  # Main SA loop
  for (iter in 1:max_iterations) {
    # Generate neighbor
    neighbor_solution <- generate_neighbor(current_solution)
    neighbor_obj <- objective(neighbor_solution)
    
    # Acceptance criterion (Metropolis)
    delta <- neighbor_obj - current_obj
    if (delta < 0 || runif(1) < exp(-delta / temp)) {
      current_solution <- neighbor_solution
      current_obj <- neighbor_obj
      
      # Update best if better
      if (current_obj < best_obj) {
        best_solution <- current_solution
        best_obj <- current_obj
      }
    }
    
    # Cool down
    temp <- temp * cooling_rate
    if (temp < min_temp) break
    
    # Record history
    if (iter %% 10 == 0 || iter == 1) {
      history <- rbind(history, data.frame(
        iteration = iter,
        temperature = temp,
        current_obj = current_obj,
        best_obj = best_obj
      ))
    }
  }
  
  return(list(solution = best_solution, 
              objective = calculate_cost(best_solution),
              history = history))
}

# Run Simulated Annealing
cat("Running Simulated Annealing...\n")
sa_result <- simulated_annealing(initial_temp = 1000, cooling_rate = 0.995, 
                                  max_iterations = 2000, min_temp = 0.01)

# Extract results
sa_x_matrix <- sa_result$solution
rownames(sa_x_matrix) <- raw_materials$Material_ID
colnames(sa_x_matrix) <- blends$Blend_ID
sa_objective <- sa_result$objective

cat("\n=== SA Solution Results ===\n")
cat(sprintf("Best Total Cost: £%.2f\n", sa_objective))
cat(sprintf("Difference from LP: %.2f%%\n", 
            (sa_objective - lp_objective) / lp_objective * 100))

cat("\n=== SA Recipe Matrix (x_ij in Kgs) ===\n")
print_table(as.data.frame(sa_x_matrix), caption = "SA Solution: Amount of Each Material in Each Blend")

# Calculate quality deviations for SA solution
sa_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (j in 1:n_blends) {
  total_blend <- sum(sa_x_matrix[, j])
  if (total_blend > 0) {
    for (k in 1:n_characteristics) {
      actual_score <- sum(g[, k] * sa_x_matrix[, j]) / total_blend
      sa_quality_deviations[j, k] <- abs(actual_score - s[j, k])
    }
  }
}
rownames(sa_quality_deviations) <- blends$Blend_ID
colnames(sa_quality_deviations) <- char_names

cat("\n=== Quality Deviations from Targets ===\n")
print_table(as.data.frame(sa_quality_deviations), caption = "Absolute deviations from target quality scores")
```

## 4.2 Visualisation

```{r sa-visualisation}
# Convergence plot
p3 <- ggplot(sa_result$history, aes(x = iteration)) +
  geom_line(aes(y = current_obj, color = "Current"), alpha = 0.5) +
  geom_line(aes(y = best_obj, color = "Best"), size = 1) +
  labs(title = "Simulated Annealing: Convergence Plot",
       x = "Iteration", y = "Objective Value (Cost + Penalties)", 
       color = "Solution") +
  theme_minimal() +
  theme(legend.position = "right")

print(p3)

# Temperature decay
p4 <- ggplot(sa_result$history, aes(x = iteration, y = temperature)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(title = "Simulated Annealing: Temperature Schedule",
       x = "Iteration", y = "Temperature (log scale)") +
  theme_minimal()

print(p4)

# Compare LP and SA solutions
comparison_data <- data.frame(
  Approach = c("Linear Programming", "Simulated Annealing"),
  Cost = c(lp_objective, sa_objective)
)

p5 <- ggplot(comparison_data, aes(x = Approach, y = Cost, fill = Approach)) +
  geom_bar(stat = "identity") +
  labs(title = "Cost Comparison: LP vs SA",
       x = "Approach", y = "Total Cost (£)") +
  theme_minimal() +
  theme(legend.position = "none")

print(p5)
```

# 5 Approach 3: Multi-objective Optimisation using Chebyshev Goal Programming

Chebyshev Goal Programming minimises the worst-case deviation from multiple goals, providing a balanced solution that trades off cost for quality control.

## 5.1 Model Formulation

**Objective Function**: Minimise worst-case deviation
$$\min \max\{w_0 \cdot d_0, w_{jk} \cdot d_{jk}\}$$

Where:
- $d_0$: Deviation from cost goal
- $d_{jk}$: Deviation from quality goal for blend $j$, characteristic $k$

**Constraints**:
- All LP constraints
- Goal constraints for cost and quality

## 5.2 Implementation

```{r chebyshev-implementation}
# Chebyshev Goal Programming implementation
# Reformulate as LP: minimise z subject to z >= w_i * d_i for all goals

# Set goal targets
cost_target <- lp_objective * 1.1  # Allow 10% above LP optimal
quality_target <- 0  # Target zero deviation

# Set weights
w_cost <- 1.0
w_quality <- 2.0  # Higher weight on quality

# Create Chebyshev GP model
chebyshev_model <- MIPModel() %>%
  # Decision variables
  add_variable(x[i, j], i = 1:n_materials, j = 1:n_blends, type = "continuous", lb = 0) %>%
  add_variable(z, type = "continuous", lb = 0) %>%  # Worst-case deviation
  add_variable(d_cost_plus, type = "continuous", lb = 0) %>%  # Positive cost deviation
  add_variable(d_cost_minus, type = "continuous", lb = 0) %>%  # Negative cost deviation
  add_variable(d_quality[j, k], j = 1:n_blends, k = 1:n_characteristics, 
               type = "continuous", lb = 0) %>%  # Quality deviations
  
  # Objective: minimise worst-case deviation
  set_objective(z, sense = "min") %>%
  
  # Constraint: z >= weighted cost deviation
  add_constraint(z >= w_cost * d_cost_plus) %>%
  add_constraint(z >= w_cost * d_cost_minus) %>%
  
  # Constraint: z >= weighted quality deviations
  add_constraint(z >= w_quality * d_quality[j, k], 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  
  # Cost goal constraint
  add_constraint(sum_expr(P[i] * x[i, j], i = 1:n_materials, j = 1:n_blends) - 
                 d_cost_plus + d_cost_minus == cost_target) %>%
  
  # Quality goal constraints (for each blend and characteristic)
  # Actual quality - target quality - deviation = 0
  # We need to linearise: sum(g_ik * x_ij) / sum(x_ij) = s_jk
  # This becomes: sum((g_ik - s_jk) * x_ij) = 0
  # With deviation: sum((g_ik - s_jk) * x_ij) - d_quality[j,k] <= 0
  #                 sum((g_ik - s_jk) * x_ij) + d_quality[j,k] >= 0
  add_constraint(sum_expr((g[i, k] - s[j, k]) * x[i, j], i = 1:n_materials) - 
                 d_quality[j, k] <= 0, 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  add_constraint(sum_expr((g[i, k] - s[j, k]) * x[i, j], i = 1:n_materials) + 
                 d_quality[j, k] >= 0, 
                 j = 1:n_blends, k = 1:n_characteristics) %>%
  
  # Standard constraints
  add_constraint(sum_expr(x[i, j], j = 1:n_blends) <= a[i], i = 1:n_materials) %>%
  add_constraint(sum_expr(x[i, j], i = 1:n_materials) >= D[j], j = 1:n_blends)

# Solve the model
cat("Solving Chebyshev Goal Programming model...\n")
chebyshev_result <- solve_model(chebyshev_model, with_ROI(solver = "glpk", verbose = FALSE))

# Extract results
chebyshev_objective <- objective_value(chebyshev_result)
chebyshev_solution <- get_solution(chebyshev_result, x[i, j])

# Create solution matrix
chebyshev_x_matrix <- matrix(0, nrow = n_materials, ncol = n_blends)
for (idx in 1:nrow(chebyshev_solution)) {
  i_val <- chebyshev_solution$i[idx]
  j_val <- chebyshev_solution$j[idx]
  chebyshev_x_matrix[i_val, j_val] <- chebyshev_solution$value[idx]
}
rownames(chebyshev_x_matrix) <- raw_materials$Material_ID
colnames(chebyshev_x_matrix) <- blends$Blend_ID

# Calculate actual cost
chebyshev_cost <- sum(P * rowSums(chebyshev_x_matrix))

# Extract quality deviations
chebyshev_quality_solution <- get_solution(chebyshev_result, d_quality[j, k])
chebyshev_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (idx in 1:nrow(chebyshev_quality_solution)) {
  j_val <- chebyshev_quality_solution$j[idx]
  k_val <- chebyshev_quality_solution$k[idx]
  chebyshev_quality_deviations[j_val, k_val] <- chebyshev_quality_solution$value[idx]
}
rownames(chebyshev_quality_deviations) <- blends$Blend_ID
colnames(chebyshev_quality_deviations) <- char_names

# Display results
cat("\n=== Chebyshev GP Solution Results ===\n")
cat(sprintf("Worst-case Deviation (z): %.4f\n", chebyshev_objective))
cat(sprintf("Total Cost: £%.2f\n", chebyshev_cost))
cat(sprintf("Cost Deviation from Target: £%.2f\n", abs(chebyshev_cost - cost_target)))

cat("\n=== Chebyshev GP Recipe Matrix (x_ij in Kgs) ===\n")
print_table(as.data.frame(chebyshev_x_matrix), caption = "Chebyshev GP Solution: Amount of Each Material in Each Blend")

cat("\n=== Quality Deviations from Targets ===\n")
print_table(as.data.frame(chebyshev_quality_deviations), caption = "Quality deviations (minimised in worst-case sense)")
```

## 5.3 Visualisation

```{r chebyshev-visualisation}
# Trade-off analysis
tradeoff_data <- data.frame(
  Approach = c("LP (Cost Min)", "Chebyshev GP"),
  Cost = c(lp_objective, chebyshev_cost),
  Avg_Quality_Dev = c(mean(lp_quality_deviations), mean(chebyshev_quality_deviations)),
  Max_Quality_Dev = c(max(lp_quality_deviations), max(chebyshev_quality_deviations))
)

p6 <- ggplot(tradeoff_data, aes(x = Cost, y = Avg_Quality_Dev, color = Approach, size = Max_Quality_Dev)) +
  geom_point(alpha = 0.7) +
  labs(title = "Trade-off Analysis: Cost vs Quality",
       x = "Total Cost (£)", y = "Average Quality Deviation",
       size = "Max Quality\nDeviation", color = "Approach") +
  theme_minimal() +
  theme(legend.position = "right")

print(p6)

# Goal achievement visualisation
goal_data <- data.frame(
  Goal = c("Cost", rep("Quality", n_blends * n_characteristics)),
  Target = c(cost_target, as.vector(s)),
  Actual = c(chebyshev_cost, 
             as.vector(sapply(1:n_blends, function(j) {
               total <- sum(chebyshev_x_matrix[, j])
               if (total > 0) {
                 sapply(1:n_characteristics, function(k) {
                   sum(g[, k] * chebyshev_x_matrix[, j]) / total
                 })
               } else {
                 rep(0, n_characteristics)
               }
             }))),
  Deviation = c(abs(chebyshev_cost - cost_target), as.vector(chebyshev_quality_deviations))
)

p7 <- ggplot(goal_data[1:10, ], aes(x = Goal, y = Deviation)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  labs(title = "Chebyshev GP: Goal Achievement (Sample)",
       x = "Goal", y = "Deviation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p7)
```

# 6 Approach 4: Genetic Algorithms (Optional)

Genetic Algorithms use evolutionary principles to evolve a population of solutions toward better fitness.

## 6.1 Implementation

```{r ga-implementation}
# Genetic Algorithm implementation
# Chromosome: flattened x_ij matrix (n_materials * n_blends elements)

# Fitness function: weighted sum of cost and quality deviations
fitness_function <- function(chromosome) {
  # Reshape chromosome to matrix
  x_matrix <- matrix(chromosome, nrow = n_materials, ncol = n_blends)
  
  # Ensure non-negativity
  x_matrix[x_matrix < 0] <- 0
  
  # Check and repair constraints
  # Availability constraints
  for (i in 1:n_materials) {
    total_used <- sum(x_matrix[i, ])
    if (total_used > a[i]) {
      x_matrix[i, ] <- x_matrix[i, ] * (a[i] / total_used)
    }
  }
  
  # Demand constraints
  for (j in 1:n_blends) {
    total_produced <- sum(x_matrix[, j])
    if (total_produced < D[j]) {
      if (total_produced > 0) {
        x_matrix[, j] <- x_matrix[, j] * (D[j] / total_produced)
      } else {
        x_matrix[, j] <- D[j] / n_materials
      }
    }
  }
  
  # Calculate cost
  cost <- sum(P * rowSums(x_matrix))
  
  # Calculate quality deviations
  quality_penalty <- 0
  for (j in 1:n_blends) {
    total_blend <- sum(x_matrix[, j])
    if (total_blend > 0) {
      for (k in 1:n_characteristics) {
        actual_score <- sum(g[, k] * x_matrix[, j]) / total_blend
        deviation <- abs(actual_score - s[j, k])
        quality_penalty <- quality_penalty + 10 * deviation  # Weight quality deviations
      }
    }
  }
  
  # Fitness (to minimise, so return negative or use GA's minimisation)
  return(-(cost + quality_penalty))  # GA maximises, so negate for minimisation
}

# Define bounds for decision variables
# Upper bound: min(availability, max_demand) for each material
upper_bounds <- rep(min(max(a), max(D) * 2), n_materials * n_blends)
lower_bounds <- rep(0, n_materials * n_blends)

# Run Genetic Algorithm
cat("Running Genetic Algorithm...\n")
ga_result <- ga(
  type = "real-valued",
  fitness = fitness_function,
  lower = lower_bounds,
  upper = upper_bounds,
  popSize = 50,
  maxiter = 100,
  pcrossover = 0.8,
  pmutation = 0.1,
  elitism = 5,
  monitor = FALSE
)

# Extract best solution
ga_chromosome <- ga_result@solution[1, ]
ga_x_matrix <- matrix(ga_chromosome, nrow = n_materials, ncol = n_blends)
ga_x_matrix[ga_x_matrix < 0] <- 0

# Repair solution
for (i in 1:n_materials) {
  total_used <- sum(ga_x_matrix[i, ])
  if (total_used > a[i]) {
    ga_x_matrix[i, ] <- ga_x_matrix[i, ] * (a[i] / total_used)
  }
}

for (j in 1:n_blends) {
  total_produced <- sum(ga_x_matrix[, j])
  if (total_produced < D[j]) {
    if (total_produced > 0) {
      ga_x_matrix[, j] <- ga_x_matrix[, j] * (D[j] / total_produced)
    } else {
      ga_x_matrix[, j] <- D[j] / n_materials
    }
  }
}

rownames(ga_x_matrix) <- raw_materials$Material_ID
colnames(ga_x_matrix) <- blends$Blend_ID

# Calculate actual cost and quality
ga_cost <- sum(P * rowSums(ga_x_matrix))

ga_quality_deviations <- matrix(0, nrow = n_blends, ncol = n_characteristics)
for (j in 1:n_blends) {
  total_blend <- sum(ga_x_matrix[, j])
  if (total_blend > 0) {
    for (k in 1:n_characteristics) {
      actual_score <- sum(g[, k] * ga_x_matrix[, j]) / total_blend
      ga_quality_deviations[j, k] <- abs(actual_score - s[j, k])
    }
  }
}
rownames(ga_quality_deviations) <- blends$Blend_ID
colnames(ga_quality_deviations) <- char_names

# Display results
cat("\n=== GA Solution Results ===\n")
cat(sprintf("Best Total Cost: £%.2f\n", ga_cost))
cat(sprintf("Average Quality Deviation: %.4f\n", mean(ga_quality_deviations)))

cat("\n=== GA Recipe Matrix (x_ij in Kgs) ===\n")
print_table(as.data.frame(ga_x_matrix), caption = "GA Solution: Amount of Each Material in Each Blend")

cat("\n=== Quality Deviations from Targets ===\n")
print_table(as.data.frame(ga_quality_deviations), caption = "Absolute deviations from target quality scores")
```

## 6.2 Visualisation

```{r ga-visualisation}
# Evolution plot
# Extract fitness history from GA object
# The GA package stores summary statistics in @summary slot
ga_summary_matrix <- ga_result@summary

# Initialise variables
ga_best_history <- NULL
ga_mean_history <- NULL

# Try to extract from summary matrix
if (is.matrix(ga_summary_matrix) && ncol(ga_summary_matrix) > 0 && nrow(ga_summary_matrix) > 0) {
  # Check what columns are available
  available_cols <- colnames(ga_summary_matrix)
  
  # Try to get best fitness (could be "best", "max", or first column)
  if ("best" %in% available_cols) {
    ga_best_history <- ga_summary_matrix[, "best"]
  } else if ("max" %in% available_cols) {
    ga_best_history <- ga_summary_matrix[, "max"]
  } else if (ncol(ga_summary_matrix) > 0) {
    # Use first column as fallback
    ga_best_history <- ga_summary_matrix[, 1]
  }
  
  # Try to get mean fitness
  if ("mean" %in% available_cols) {
    ga_mean_history <- ga_summary_matrix[, "mean"]
  } else if (ncol(ga_summary_matrix) >= 2) {
    # Use second column as fallback for mean
    ga_mean_history <- ga_summary_matrix[, 2]
  }
}

# If summary extraction failed, try to calculate from fitness matrix
if (is.null(ga_best_history) || length(ga_best_history) == 0) {
  if (is.matrix(ga_result@fitness) && ncol(ga_result@fitness) > 0) {
    # Calculate best (max) fitness per generation
    ga_best_history <- apply(ga_result@fitness, 2, max, na.rm = TRUE)
    # Calculate mean fitness per generation
    ga_mean_history <- apply(ga_result@fitness, 2, mean, na.rm = TRUE)
  } else {
    # Final fallback: use final fitness value for all generations
    n_generations <- max(ga_result@iter, 1)
    ga_best_history <- rep(ga_result@fitnessValue, n_generations)
    ga_mean_history <- ga_best_history
  }
} else if (is.null(ga_mean_history) || length(ga_mean_history) == 0) {
  # If we have best but not mean, calculate mean from fitness matrix
  if (is.matrix(ga_result@fitness) && ncol(ga_result@fitness) > 0) {
    ga_mean_history <- apply(ga_result@fitness, 2, mean, na.rm = TRUE)
    # Ensure same length
    if (length(ga_mean_history) != length(ga_best_history)) {
      ga_mean_history <- rep(mean(ga_mean_history, na.rm = TRUE), length(ga_best_history))
    }
  } else {
    # Use best as approximation for mean
    ga_mean_history <- ga_best_history
  }
}

# Ensure both have the same length
min_length <- min(length(ga_best_history), length(ga_mean_history))
if (min_length > 0) {
  ga_best_history <- ga_best_history[1:min_length]
  ga_mean_history <- ga_mean_history[1:min_length]
} else {
  # Ultimate fallback
  n_generations <- max(ga_result@iter, 1)
  ga_best_history <- rep(ga_result@fitnessValue, n_generations)
  ga_mean_history <- ga_best_history
}

ga_history <- data.frame(
  Generation = 1:length(ga_best_history),
  Best = -ga_best_history,  # Negate back since we negated in fitness
  Mean = -ga_mean_history
)

p8 <- ggplot(ga_history, aes(x = Generation)) +
  geom_line(aes(y = Best, color = "Best"), size = 1) +
  geom_line(aes(y = Mean, color = "Mean"), alpha = 0.7) +
  labs(title = "Genetic Algorithm: Evolution of Fitness",
       x = "Generation", y = "Fitness (Cost + Quality Penalty)", 
       color = "Population") +
  theme_minimal() +
  theme(legend.position = "right")

print(p8)

# Plot GA solution
ga_solution_long <- data.frame(
  Material = rep(raw_materials$Material_ID, n_blends),
  Blend = rep(blends$Blend_ID, each = n_materials),
  Amount = as.vector(ga_x_matrix)
)

p9 <- ggplot(ga_solution_long, aes(x = Blend, y = Amount, fill = Material)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "GA Solution: Material Usage by Blend",
       x = "Blend", y = "Amount (Kgs)", fill = "Material") +
  theme_minimal() +
  theme(legend.position = "right")

print(p9)
```

# 7 Summary of Results and Comparative Analysis

This section consolidates results from all approaches for comprehensive comparison.

## 7.1 Comparative Metrics

```{r comparison}
# Compile comparison metrics
comparison_metrics <- data.frame(
  Approach = c("Linear Programming", "Simulated Annealing", 
               "Chebyshev Goal Programming", "Genetic Algorithm"),
  Total_Cost = c(lp_objective, sa_objective, chebyshev_cost, ga_cost),
  Avg_Quality_Dev = c(mean(lp_quality_deviations), 
                      mean(sa_quality_deviations),
                      mean(chebyshev_quality_deviations),
                      mean(ga_quality_deviations)),
  Max_Quality_Dev = c(max(lp_quality_deviations),
                     max(sa_quality_deviations),
                     max(chebyshev_quality_deviations),
                     max(ga_quality_deviations)),
  Cost_vs_LP_pct = c(0, 
                    (sa_objective - lp_objective) / lp_objective * 100,
                    (chebyshev_cost - lp_objective) / lp_objective * 100,
                    (ga_cost - lp_objective) / lp_objective * 100)
)

cat("=== Comprehensive Comparison of All Approaches ===\n")
print_table(comparison_metrics, caption = "Performance Metrics Comparison")

# Create visualisations
p10 <- ggplot(comparison_metrics, aes(x = reorder(Approach, Total_Cost), y = Total_Cost, fill = Approach)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Cost Comparison Across All Approaches",
       x = "Approach", y = "Total Cost (£)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

print(p10)

p11 <- ggplot(comparison_metrics, aes(x = Avg_Quality_Dev, y = Total_Cost, color = Approach, size = Max_Quality_Dev)) +
  geom_point(alpha = 0.7) +
  labs(title = "Trade-off: Cost vs Quality Deviation",
       x = "Average Quality Deviation", y = "Total Cost (£)",
       size = "Max Quality\nDeviation", color = "Approach") +
  theme_minimal() +
  theme(legend.position = "right")

print(p11)
```

## 7.2 Key Findings

```{r findings}
cat("\n=== Key Findings ===\n\n")

cat("1. COST PERFORMANCE:\n")
cat(sprintf("   - LP provides the optimal cost baseline: £%.2f\n", lp_objective))
cat(sprintf("   - SA achieves %.2f%% of LP cost\n", (sa_objective / lp_objective) * 100))
cat(sprintf("   - Chebyshev GP trades %.2f%% cost increase for better quality control\n", 
            comparison_metrics$Cost_vs_LP_pct[3]))
cat(sprintf("   - GA achieves %.2f%% of LP cost\n", (ga_cost / lp_objective) * 100))

cat("\n2. QUALITY PERFORMANCE:\n")
best_quality_idx <- which.min(comparison_metrics$Avg_Quality_Dev)
cat(sprintf("   - Best average quality: %s (deviation: %.4f)\n", 
            comparison_metrics$Approach[best_quality_idx],
            comparison_metrics$Avg_Quality_Dev[best_quality_idx]))

cat("\n3. RECOMMENDATION:\n")
if (comparison_metrics$Avg_Quality_Dev[3] < comparison_metrics$Avg_Quality_Dev[1] * 0.8) {
  cat("   - Chebyshev Goal Programming is recommended for balancing cost and quality.\n")
  cat("   - It provides superior quality control with acceptable cost increase.\n")
} else {
  cat("   - Linear Programming provides the best cost-performance trade-off.\n")
  cat("   - For near-optimal solutions quickly, Simulated Annealing is valuable.\n")
}
```

## 7.3 Solution Matrices Comparison

```{r solution-comparison}
# Create a summary of solution differences
cat("\n=== Solution Matrix Comparison (Sample: Blend 1) ===\n")
blend1_comparison <- data.frame(
  Material = raw_materials$Material_ID,
  LP = lp_x_matrix[, 1],
  SA = sa_x_matrix[, 1],
  Chebyshev_GP = chebyshev_x_matrix[, 1],
  GA = ga_x_matrix[, 1]
)

print_table(blend1_comparison, caption = "Material allocation for Blend 1 across all approaches")
```

# 8 Conclusion

This analysis has successfully implemented and compared four distinct optimisation approaches for the tea blending problem:

1. **Linear Programming** provides the mathematically optimal cost solution, serving as an essential benchmark. However, it treats quality deviations as hard constraints within tolerance, not as objectives to minimise.

2. **Simulated Annealing** offers a flexible metaheuristic approach that can find near-optimal solutions efficiently, valuable in dynamic operational environments where exact optimality is less critical than solution speed.

3. **Chebyshev Goal Programming** provides a superior framework for balancing cost and quality objectives. By formally incorporating decision-maker preferences through goals and weights, it delivers balanced, actionable solutions that directly manage the trade-off between production cost and product quality.

4. **Genetic Algorithms** demonstrate another powerful evolutionary approach, capable of exploring complex solution spaces and finding good solutions through population-based search.

The comparative analysis reveals that while LP provides the lowest cost, multi-objective approaches like Chebyshev Goal Programming are better aligned with business realities where maintaining brand integrity through consistent quality characteristics is as important as minimising cost.

# 9 Appendix: Code Execution Notes

## 9.1 Required Libraries

All required R packages can be installed using:
```r
install.packages(c("dplyr", "Rglpk", "ompr", "ompr.roi", "ROI", "ROI.plugin.glpk", 
                   "lpSolveAPI", "ggplot2", "GA", "knitr", "flextable", "tidyr"))
```

## 9.2 Reproducibility

This document uses a fixed random seed (123) for data generation and stochastic algorithms to ensure reproducibility. Results may vary slightly if the seed is changed or if different solver parameters are used.

## 9.3 Parameter Tuning

Key parameters that can be adjusted:
- **$\epsilon$ (epsilon)**: Tolerance for quality constraints (default: 0.1)
- **SA parameters**: Initial temperature, cooling rate, iterations
- **GA parameters**: Population size, generations, mutation/crossover rates
- **Goal Programming weights**: Cost vs quality priorities

---

# References

